Intersecting experimental evolution and CRISPR screens to identify novel toxin resistance loci

Version 1

Abstract

Understanding toxin resistance in insects is key to appreciate niche adaptations but remains challenging due to its often-polygenic basis. A well-known example is the specialized association of Drosophila sechellia with noni fruit (Morinda citrifolia), which is toxic to most other insects, including the closely-related drosophilids D. simulans and D. melanogaster. Noni toxicity is due to its high concentration of octanoic acid (OA), but the mechanisms that determine sensitivity or resistance to OA remain poorly understood. Here, we experimentally-evolved D. simulans with increased OA resistance, identifying multiple loci under selection. Cross-referencing these with a genome-wide, OA-resistance CRISPR screen in a D. melanogaster cell line highlighted two proteins: Kraken, a putative detoxification enzyme expressed in digestive and renal tissues, and Alkbh7, a mitochondrial protein linked to fatty acid metabolism. Both genes show elevated expression in D. sechellia and OA-resistant D. simulans. In D. melanogaster, kraken mutants are more OA-sensitive, while Alkbh7 overexpression increased OA resistance. Importantly, mutation of these genes in D. sechellia reduced OA tolerance. Our identification of genes underlying OA resistance in laboratory and natural contexts demonstrates how complementary, cross-species selection approaches can provide insights into complex mechanisms of toxin susceptibility and adaptation, and have practical applications in the characterization of novel insecticides.

Introduction

How insects resist toxic chemicals in the environment is of both basic scientific interest – reflecting insects’ arms race with plant food sources (Erb and Reymond 2019; Jones, et al. 2022) or venomous predators (Smith, et al. 2013) – and applied interest, because, throughout history, humans have used chemicals to combat insect disease vectors and agricultural pests (Umetsu and Shirai 2020). The best-understood insecticides are those that bind to and modulate neuronal ion channels, where resistance arises through changes in these target proteins (Sparks, et al. 2020; Raisch and Raunser 2023). For example, dichlorodiphenyltrichloroethane (DDT) and pyrethroids prevent closure of voltage-gated sodium channels, resulting in neuronal excitotoxicity and organismal paralysis; diverse insect species displaying resistance for these chemicals have independently-arising mutations in genes encoding these channels (Busvine 1951; Dong 2007). Similarly, neonicotinoids bind to and hyperactivate nicotinic acetylcholine receptors, leading to neuronal death; mutations in various receptor subunits can confer resistance (Matsuda, et al. 2020). Even in such well-studied examples, however, it is thought that other neuronal targets exist (Soderlund and Bloomquist 1989) and/or that the chemicals impact receptors in non-neuronal tissues (Di Prisco, et al. 2013). For many insecticides – as well as other environmental pollutants (Gandara, et al. 2024) – the mode of action is incompletely, or not at all, characterized (Sparks, et al. 2020).

Resistance observed in nature is only partially understood, reflecting the contribution of many genes beyond those encoding target proteins, which can underlie distinct defence mechanisms (Clarkson, et al. 2018; Catteruccia 2020; Lucas, et al. 2023). For example, thickening of the cuticle to reduce entry of insecticides into the body is widely considered an effective resistance mechanism (Balabanidou, et al. 2018; Balabanidou, et al. 2019). Cuticular changes have been linked to overexpression of proteins implicated in exoskeleton development (e.g., cuticle proteins, ABC transporters) but their precise roles, and how they are upregulated, remain unclear (Balabanidou, et al. 2018). Increased expression of metabolic enzymes that degrade insecticides is another commonly observed adaptation in resistant strains, sometimes resulting from copy-number increases in the corresponding genes (Weetman, et al. 2018). Duplicated metabolic genes can also undergo neofunctionalization to produce enzymes with altered specificity, as illustrated by Cytochrome P450s in strains of the brown planthopper that recently evolved resistance to the neonicotinoid imidacloprid (Zimmer, et al. 2018), as well as by more ancient evolution of glutathione S-transferases during transition of the Scaptomyza genus of Drosophilidae to herbivory to counter toxic isothiocyanates (Gloss, et al. 2014). However, the lack of experimental resources in many insect species renders it difficult to characterize the causal contribution of specific genetic changes.

A compelling natural model clade to study the evolution of toxin resistance is the trio of closely-related drosophilid species, Drosophila melanogaster, Drosophila simulans and Drosophila sechellia (Fig. 1A). D. sechellia is an extreme specialist on the toxic noni fruit of the Morinda citrifolia shrub (Jones 2005; Auer, et al. 2021). Amongst D. sechellia’s numerous adaptations (R’Kha, et al. 1991; Dekker, et al. 2006; Auer, et al. 2020; Auer, et al. 2021; Alvarez-Ocana, et al. 2023; Shahandeh, et al. 2024), this species has evolved high resistance to octanoic acid (OA), an abundant noni chemical that is toxic to other drosophilids and more divergent insects (Legal, et al. 1994; Jones 1998; Legal, et al. 1999; Markow 2019; Auer, et al. 2021; Kaczmarek, et al. 2022; Kaczmarek, et al. 2024). Previous pioneering studies of D. sechellia’s resistance to OA has been studied through genomic mapping-based approaches in D. sechellia/D. simulans hybrids, which revealed this trait to be polygenic, comprising at least five genomic regions (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015). One of the main effect loci encompasses a cluster of 18 genes including members of the Osiris (Osi) family, which are thought to function in cuticle patterning (Ando, et al. 2019; Sun, et al. 2024). RNA interference (RNAi) knockdown of some Osi genes in D. melanogaster led to reduced tolerance of OA (Andrade Lopez, et al. 2017; Lanno, et al. 2019). However, genetic evidence for the contribution of these genes to OA resistance in D. sechellia is lacking. Given the challenges of identifying candidate genes through genomic mapping-based strategies, in this work we took two complementary approaches to investigate the evolution of OA resistance in this drosophilid trio: experimental evolution in D. sechellia’s closest relative, D. simulans, through an Evolve-and-Resequence strategy (Schlotterer, et al. 2016), and CRISPR/Cas9-based genome-wide genetic screening in cultured D. melanogaster cells (Viswanatha, et al. 2018). We show how the intersection of these approaches can identify novel genes contributing to OA susceptibility/resistance in these drosophilids, establishing these methods and this species clade as a powerful paradigm to investigate ecotoxicology.

Experimental evolution of D. simulans with increased OA resistance.

A Phylogeny of the drosophilid trio studied in this work. Mya: million years ago. B Schematic of octanoic acid (OA) resistance tube assay (see Materials and Methods) and selection procedure. This panel was created using BioRender.com. C Species differences in resistance of D. simulans 04 and D. sechellia 07 strains to 3 μl OA. D Schematic illustrating the generation of the base populations for Evolve-and-Resequence (see Materials and Methods). This panel was created using BioRender.com. E Changes in survivability to 3 μl OA of the ten D. simulans populations over generations 0-25. Box plots represent the interquartile ranges of the survivability for all the replicates; blue lines shows the generalized additive model (GAM) regression; grey shading represents the 95% confidence interval for the mean responses at each time point. F As in E for the 10 populations combined. G OA resistance for the combined 10 populations with increasing OA volumes during generation 25-31. H As in E, with 9 μl OA during generations 31-50. I As in H for the 10 populations combined. J Comparison of OA resistance of unevolved D. simulans (i.e., the original 10 outcrossed populations maintained on standard fly food; see Materials and Methods), evolved (generation 50 (G50)) D. simulans, and D. sechellia. K-N Phenotypic comparison of unevolved and evolved (G50+) D. simulans populations 1-3 for K noni resistance (G54), L fecundity (G54), M developmental viability (G57), and N lifespan (G57). For K-M, significance was assessed using the unpaired t-test correcting for multiple testing. NS p>0.05, ***p<0.001. For L-N, tests were run on standard media. No significant differences between unevolved and evolved flies were detected using the Cox regression model. See Materials and Methods for details and Data S1 for raw data.

Results

Experimental evolution of OA resistance in D. simulans

A previous study demonstrated that D. simulans strains can be experimentally evolved for higher OA resistance, enabling the identification of genomic regions (through microsatellite markers) responding to selection (Colson 2004). This success inspired us to take an Evolve-and-Resequence approach, which uses high-throughput sequencing to analyse the entire genome of populations before and after exposure to specific selective pressures, thereby facilitating the identification of adaptive loci from standing genetic variation in controlled and replicable settings (Turner and Miller 2012; Schlotterer, et al. 2016; Barghi, et al. 2019; Kelly and Hughes 2019; Shahrestani, et al. 2021). For studies of insecticide resistance, it has been predominantly used to examine changes in frequency of alleles of known genes rather than as a means to identify novel loci (Zoh, et al. 2021; Sadia, et al. 2024).

The susceptibility/resistance of drosophilids to OA can be easily assessed by exposing flies to the toxin in a culture tube and counting the survivors after 24 h (Fig. 1B). In line with previous observations (Amlou, et al. 1997; Colson 2004), we found that a dose of 3 μl of pure OA (mixed with 1 ml glucose solution), corresponding to roughly half the amount estimated to be present in 2 g of noni pulp (see Materials and Methods), leads to death of ∼60% of D. simulans but has no effect on D. sechellia (Fig. 1C). By collecting the D. simulans survivors, we reasoned that we could allow these to mate and establish the next generation (Fig. 1B), thus enriching genetic variants conferring higher resistance, similarly to (Colson 2004). Repeating the process over multiple generations and multiple replicates would allow us to track phenotypic changes in resistance and investigate the associated trajectories in allele frequencies.

Standing genetic variation should be present in the evolving populations for natural selection to act upon (Teotonio, et al. 2009; Long, et al. 2015). As the spontaneous mutation rate in drosophilids is too low for artificial selection (Keightley, et al. 2014), we generated ten independent polymorphic base populations of D. simulans from 93 isogenic strains (Signor, et al. 2018) by round-robin crossing (Fig. 1D). These lines contain millions of SNPs, but we note that they were sampled from a single wild population in California, therefore reflecting the genetic variation present in that locality. The base populations therefore might not include all variants found across the species’ range, including those in regions where D. sechellia co-occurs. To minimize bottleneck effects, each population was kept at a size of ∼1000 flies. At each generation, ∼500 flies from each population were randomly collected and selected for resistance to 3 μl OA. Over the course of 25 generations, all the populations showed increased levels of survivability, suggesting parallel adaptive evolution (Fig. 1E). Given the consistent responses across all populations, we also analysed the phenotypic changes grouping all replicates, finding that the mean survivability significantly increased from ∼43% at generation 0 to ∼78% at generation 25 (Fig. 1F).

As we kept the same amount of OA throughout the 25 generations of selection, the selective pressure lowered as the populations’ survivability increased, evidenced as a larger fraction of flies surviving and reproducing. Viewing the phenotype of populations both individually (Fig. 1E) and collectively (Fig. 1F), it was apparent that the change in phenotype plateaued at around generation 15. For continued evolution, we therefore gradually increased the amount of OA during generations 26-31 (Fig. 1G), reaching a selection pressure comparable to the initial phase of the experiment with 9 μl OA (∼38% mean survivability across all populations) (Fig. 1G). We then performed a second round of selection at this dose, observing an increase in survivability across all populations, reaching ∼70% by generation 50 (Fig. 1H). As the populations adapted to the new condition, a second plateau started around generation 40 (Fig. 1H-I). Over 50 generations we were therefore able to evolve D. simulans to have approximately 5-fold higher survivability to 9 μl OA compared to the unevolved populations (Fig. 1J), though still falling short of the survival exhibited by D. sechellia (Fig. 1J). These plateaus can be explained by the decreasing selective pressure exerted by fixed OA volumes as an ever-greater proportion of flies adapt and survive the treatment. Determining whether an upper limit exists that would constrain the evolution of a D. sechellia-like level of phenotypic resistance would require additional selection rounds using progressively higher OA concentrations over an extended (and unknown) number of generations. OA is only one component of noni fruit, although probably the major toxic chemical (Legal, et al. 1994). We therefore tested three of our evolved populations for resistance to noni fruit pulp, finding that they exhibited much higher survival on this substrate compared to the same original populations maintained under neutral conditions (i.e., no OA selection) (Fig. 1K).

Specialization of D. sechellia on toxic noni is associated with a reduced egg-laying capacity (R’Kha, et al. 1997; Markow, et al. 2009) and a shorter (albeit strain-specific) lifespan (Watada, et al. 2020; Abe, et al. 2022; Shahandeh, et al. 2024). Our success in evolving D. simulans with a partial D. sechellia-like resistance trait begs the question as to whether these selected strains exhibit other phenotypic changes as a trade-off for higher toxin resistance. We tested whether the evolved resistant populations of D. simulans exhibit similar phenotypic changes in control conditions. However, none of the three focal populations tested exhibited significant decreases in the number of eggs laid (Fig. 1L), developmental viability (Fig. 1M) or lifespan (Fig. 1N). Thus, at least within the timeframe of our experimental evolution experiment, D. simulans did not display obvious phenotypic trade-offs for the gain in OA resistance.

Dynamic directional selection of allele frequencies

We hypothesized that the observed phenotypic changes in D. simulans depended on the directional selection of alleles of consistent loci across the ten populations, while most of the genome evolves under neutrality. Allele frequencies at the selected loci should therefore consistently change across all populations, allowing us to statistically distinguish them from randomly evolving loci (Fig. 2A). We sequenced the genomes of the populations of evolved (generation 25 and generation 50) and unevolved (generation 0) flies, identifying 5,127,840 high-quality variants distributed across all chromosomes. Pairwise identity-by-state analysis indicated that as the number of generations increased, the proportions of shared alleles decreased, both between time points and between pairs of populations (left-to-right and top-to-bottom, respectively, in Fig. 2B).

Dynamic directional selection of allele frequencies during experimental evolution.

A Schematic of the analysis. B Identity-by-state analysis. The heatmap depicts the proportion of shared alleles between populations, with values ranging from 0.8 (blue, lowest shared proportion) to 1 (red, identical alleles). C,D Manhattan plot showing the probability of allele frequency changes between C generations 0 and 25, and D generations 0 and 50. Top panel shows the results of the CMH test; orange dots indicate significant variants at a threshold of -log10(p) ≥40, a conservative approximation of the top 0.000005% simulated SNPs under neutrality (Fig. S1B). Bottom panel shows the absolute log-Bayes factors estimated by Bait-ER; orange dots indicate significant variants at the conservative threshold log(0.99/0.01). Raw data in Data S2. E Circular plots illustrating overlap of predicted genomic regions under directional selection between the two statistical approaches at generation 25 (left) and generation 50 (middle), or between generation 25 and 50 (right). Raw data in Data S3.

To identify the selected genetic variants underlying the phenotypic adaptation, we analysed the genome-wide allele frequency changes both between generations 0 and 25, and between generations 0 and 50, across all populations. The identification of significant changes in frequency is non-trivial, and many methods have been developed (Vlachos, et al. 2019; Kojima, et al. 2020; Barata, et al. 2023). We therefore assessed the significance of frequency changes using two complementary approaches: the Cochran–Mantel–Haenszel (CMH) test (Cochran 1954; Mantel and Haenszel 1959), a widely-used method that tests for allele frequency changes between two time points, and Bait-ER (Barata, et al. 2023), a Bayesian method that tests for selection while explicitly modelling genetic drift. Both tests identified large portions of the genome under positive selection at generation 25 and 50 (Fig. 2C,D). The majority of prominent peaks were found by both methods and across time points (Fig. 2E).

Based on the analysis of genome-wide linkage disequilibrium decay (Fig. S1A), we grouped the significant SNPs identified by each of the two methods into genomic blocks according to their proximity (i.e., <50 kb). We retrieved all the candidate genes that fall within the block intervals, obtaining a total of 439 and 401 genes for the CMH and Bait-ER, respectively, at generation 25, and 1238 and 174 genes at generation 50 (Data S3). These genes likely include true targets of selection, but also neighbouring genes affected by hitchhiking. The increase in CMH candidate genes over time – in contrast to the decrease observed with Bait-ER – might reflect the CMH test’s tendency to overestimate significance and to be insensitive to divergence among replicate populations.

The overlap of all significant genes between the two time points is <30%. This might reflect that a large fraction of the candidate genes results from noise caused by genetic drift and draft (hitchhiking) and/or reflect a shift in the phenotypic contribution of genes under the stronger selection (i.e., the higher dose of OA) during generations 27-50. Moreover, previous work on Drosophila using temporal sampling showed that adaptation involving complex traits can lead to non-overlapping sets of selected SNPs, in contrast with the simple expectation of continuous increases in allele frequency until fixation (Orozco-terWengel, et al. 2012; Schlotterer, et al. 2016). Gene ontology analysis is complicated at this stage as most hits are likely hitchhikers of the true causal genes. It is also hard to compare currently to previous mapping studies in D. simulans/D. sechellia hybrids (Amlou, et al. 1997; Amlou, et al. 1998; Huang and Erezyilmaz 2015) or in the earlier D. simulans experimental evolution experiment (Colson 2004), which only identified wide genomic regions. The sole exception, a 170-kb window on 3R containing only 18 genes (including several Osiris genes) (Hungate, et al. 2013) did not emerge in our analysis.

A genome-wide, pooled CRISPR screen for OA susceptibility and resistance genes in D. melanogaster S2R+ cells

Given that these previous studies also contain extensive hitchhiking signals and substantial noise across large genomic regions, comparing their results with our analyses would likely yield numerous false positives and still leave a large pool of putative loci, offering limited benefit for fine-mapping. We therefore sought an orthogonal screening approach. Motivated by previous observations of OA toxicity for insect Sf9 cultured cells (Kaczmarek, et al. 2022), we found that OA also leads to reduced viability of D. melanogaster S2R+ cells (Fig. 3A). This phenotype prompted us to perform a selection experiment in this cell line through exploitation of a genome-wide, pooled CRISPR/Cas9 mutagenesis approach (Fig. 3B) (Viswanatha, et al. 2018). In brief, a library of sgRNA-encoding plasmids (6-8 sgRNAs/gene) was transfected into Cas9-expressing cells where site-specific plasmid integration predominantly limits targeting to a single gene per cell. Two replicate cultures of this pool of cells were either left untreated, or exposed to OA in the medium, at two different concentrations (Fig. 3B). After 26 days in culture, DNA was extracted from the cell pool, then sgRNA sequences were PCR amplified and sequenced to determine those that were enriched after OA selection (reflecting genes whose loss enhances cell viability, referred to hereafter as “susceptibility genes”) or depleted (reflecting genes whose loss leads to lower OA resistance, hereafter “resistance genes”).

A genome-wide pooled CRISPR screen for OA susceptibility and resistance genes in D. melanogaster cultured cells.

A Left: images of S2R+ cells cultured in the absence or presence of OA. Scale bar, 50 μm. Right: dose-dependent effects of OA on S2R+ cell viability, through measurement of ATP levels with a CellTiter-Glo assay (see Materials and Methods). B Schematic of CRISPR screen design. C Top 5% “susceptibility” genes (see Materials and Methods and Data S4). Box plots represent the interquartile ranges for all the replicates of the Z-scores obtained through maximum likelihood estimation to test for selection. The inset (bottom-right) displays the gene-ontology analysis of these genes. D As in C, for the top 5% “resistance” genes. E Venn diagram showing the overlap between all the genes identified in the experimental evolution at G25 and G50, and genes associated with susceptibility and resistance in the CRISPR screen. F Subportion of the top panel (CMH test) of the Manhattan plot from Fig. 2C for chromosome arms 2L and 3L, with the red arrows highlighting the regions magnified on the right showing the significant variants and linkage disequilibrium blocks (triangles) for kraken and Alkbh7 (CG14130).

To identify candidate OA susceptibility/resistance genes, we first filtered out genes essential for general cell survival (Viswanatha, et al. 2018) and genes that were not detectably expressed in S2R+ cells (using DGET (Hu, et al. 2017). (Removal of such genes are inevitable limitations of this cell-based screening approach; see Discussion). After comparing results from each condition and replicate, we identified the top 5% susceptibility genes (Fig. 3C) and top 5% resistance genes (Fig. 3D). Amongst the 87 susceptibility genes, there was no single stand-out hit, suggesting that – at least in S2R+ cells – OA does not have a single major cellular target (unlike, for example, insecticidal bacterial toxins that bind a specific cell surface receptor (Xu, et al. 2022)). Gene-ontology analysis of these susceptibility genes revealed a significant enrichment of genes involved membrane trafficking (Fig. 3C). As OA has previously been hypothesized to disrupt cellular membranes (Borrull, et al. 2015; Kaczmarek, et al. 2022), the recovery of these hits suggests that general reduction in membrane transport reduces OA uptake into cells and/or renders membranes less susceptible to OA treatment. 22 OA resistance genes were identified, encoding diverse types of proteins (Fig. 3D), suggesting there are multiple mechanisms by which cells can resist the toxic effects of OA.

We next examined which of the genes identified in the S2R+ cell selection experiments might be relevant for evolution of OA resistance in whole animals by determining those also present within significant blocks of the experimental evolution analyses (Fig. 3E,F). Four susceptibility genes were identified: CG1620 (a predicted SANT-MYB domain transcription factor), CG4291 (a putative component of the splicing machinery), Nph (a chromatin remodelling factor (Emelyanov, et al. 2014)) and Debcl (a pro-apoptotic member of the Bcl-2 family (Galindo, et al. 2009)); these are all likely to have indirect roles in conferring susceptibility to OA given their broad functions. By contrast, two resistance genes have potentially direct roles in conferring tolerance to OA, kraken and CG14130. kraken encodes a putative serine hydrolase (Edwin Chan, et al. 1998) and stood out both because it was the top hit in the resistance gene screen (Fig. 3D), and because of its location in one of the most prominent peaks in the experimental evolution at generations 25 and 50 (Fig. 3F and Fig. 2C-D). CG14130 encodes the orthologue of mammalian Alkbh7, which is involved in regulating fatty acid metabolism (Solberg, et al. 2013; Zhang, et al. 2021), and is located within a significant peak at generation 25 (Fig. 3F and Fig. 2C). Therefore, intersecting the experimental evolution and CRISPR screening results allowed us to narrow down tens or hundreds of potential candidates to two focal genes for downstream investigations.

Sequence, expression and evolutionary analyses of candidate OA resistance genes

To assess the candidacy of kraken and CG14130 (hereafter Alkbh7) in contributing to OA resistance in animals, we first examined their sequence and expression in D. melanogaster. Kraken belongs to a large family of serine hydrolases, which encompass lipases, esterases and proteases that use a nucleophilic serine residue in the active site for hydrolysis of substrates (Edwin Chan, et al. 1998). Consistent with the conservation of this serine in Kraken (Fig. 4A and Fig. S2A), an activity-based proteomic screen of the serine hydrolase superfamily in D. melanogaster provided evidence for in vivo catalytic activity of Kraken (Kumar, et al. 2021). kraken is broadly expressed in adults, with particular enrichment in the gut and Malpighian tubules (Fig. 4B), similar to previous observations of kraken expression in the embryo by RNA in situ hybridization (Edwin Chan, et al. 1998). Using the Fly Cell Atlas (Li, et al. 2022), we examined sub-tissue adult expression of kraken: in the Malpighian tubules, kraken is expressed at highest levels in the principal cells (Fig. 4C), which are involved in ion transport and detoxification. In the gut, kraken transcripts are most robustly detected in various types of enterocytes, which secrete digestive enzymes (Lemaitre and Miguel-Aliaga 2013) (Fig. 4C). However, Kraken lacks an N-terminal signal sequence (Teufel, et al. 2022), suggesting that the protein remains in the cytoplasm of these cells, unlike canonical digestive enzymes (Lemaitre and Miguel-Aliaga 2013), and more similar to detoxification enzymes of the Cytochrome P450 and Glutathione S-transferase families (Yang, et al. 2007). To examine kraken expression in vivo, we performed RNA fluorescence in situ hybridization, which detected transcripts in various region of the gut (crop, crop duct, proventriculus, hindgut, ampulla) and Malpighian tubules (Fig. 4D).

Sequence and expression analyses of kraken and Alkbh7.

A AlphaFold protein models for D. melanogaster Kraken (AF-O18391-F1-model_v4) and Alkbh7 (AF-Q9VTP1-F1-model_v4) (Jumper, et al. 2021; Varadi, et al. 2022), highlighting a conserved putative catalytic serine in the active site in Kraken, and the mitochondrial targeting sequence in Alkbh7 (predicted by MitoFates (Fukasawa, et al. 2015)). See also Fig. S2A,B. B Tissue-specific expression of D. melanogaster kraken and Alkbh7 from bulk RNA-sequencing data from the Fly Atlas 2.0 (Krause, et al. 2022). C tSNE plots illustrating D. melanogaster kraken expression in single-cell transcriptomes of the Malpighian tubules and gut from the Fly Cell Atlas (Stringent 10x datasets) (Li, et al. 2022). D Left: RNAscope detection of kraken (green) transcripts in the gut and Malpighian tubules, with nuclei counterstained with DAPI (blue). Right: higher-magnification images showing kraken transcript expression in the indicated tissues. Scale bars, 200 μm. This expression pattern was observed in tissues from >20 individuals. E Expression levels of the kraken and Alkbh7 in the indicated laboratory strains (left panel) and wild caught strains (right panel) of adult female D. sechellia and D. simulans measured by qPCR (Data S6). Expression is represented as calibrated and normalized relative quantities (CNRQ). Significance was assessed using the unpaired t-test correcting for multiple testing and represented using letter codes. F Expression levels of kraken and Alkbh7 in the indicated strains and species at larval or adult stages as indicated from published RNA-seq data. (References: 1, {Watanabe, 2019 #6929}, 2 {Ma, 2018 #8721}, 3 {Kalra, 2024 #10088}). Significance was assessed using the unpaired t-test correcting for multiple testing and represented using letter codes when multiple replicates were present in the original data. G Expression levels of the candidate genes at generation 0, 25 and 50 of the experimentally-evolved D. simulans populations measured by qPCR (Data S6). Significance was assessed using the paired t-test correcting for multiple testing and represented using letter codes.

Alkbh7 is a member of the alpha-ketoglutarate-dependent hydroxylase (Alkbh) family (Fig. 4A), which are enzymes that act on diverse substrates in processes such as biosynthesis, metabolism and epigenetic regulation (Xu, et al. 2021). As in mammals (Solberg, et al. 2013), D. melanogaster Alkbh7 displays broad tissue expression (Fig. 4B). Murine Alkbh7 is localized to the mitochondrial matrix (Solberg, et al. 2013). This property appears to be conserved in the Drosophila protein, which has a mitochondrial targeting sequence (Fig. 4A and Fig. S2B) (Fukasawa, et al. 2015); consistently, we identified Alkbh7 within D. melanogaster proteomes of the mitochondrial matrix (Chen, et al. 2015; Sen and Cox 2022).

Although the sequences of both Kraken and Alkbh7 are overall highly similar across D. melanogaster, D. simulans and D. sechellia (Fig. S2A,B), we wondered whether we could detect signs of selection at these loci in D. sechellia employing population sequence datasets of D. sechellia and D. simulans (Schrider, et al. 2018). Using the McDonald-Kreitman test (McDonald and Kreitman 1991) on the coding sequences of both genes, we found evidence for gene-level positive selection on kraken but not Alkbh7 (Fig. S2C). Because selection might act pervasively at individual sites rather than along the entire protein coding region, we also implemented a codon-based analysis using FUBAR (Fast, Unconstrained Bayesian AppRoximation) (Murrell, et al. 2013), which identified 1 and 4 codons under diversifying positive selection in kraken and Alkbh7, respectively (Fig. S2D). Although their functional relevance remains unclear, such mutations might reflect underlying evolutionary dynamics at the protein level.

We next compared the expression levels of these genes both in laboratory and wild-caught strains of D. sechellia and D. simulans (Fig. 4E), as well as in RNA-seq datasets of various strains of D. sechellia, D. simulans and D. melanogaster (Ma, et al. 2018; Watanabe, et al. 2019; Kalra, et al. 2024) (Fig. 4F, Fig. S3A,B). While there were some exceptions, overall both kraken and Alkbh7 show higher expression levels in D. sechellia adults and larvae compared to the other species. Importantly, in our experimentally-evolved D. simulans populations, we also found that both kraken and Alkbh7 were more highly expressed compared to unselected populations (Fig. 4G). The expression differences between species, and between unevolved and evolved D. simulans, reflect variation in basal levels of expression, as prior exposure to OA did not lead to changes in expression in either D. sechellia or

D. simulans (Fig. S3C). RNA FISH of kraken in D. sechellia revealed similar spatial expression as in D. melanogaster, with the addition of a robust signal in the midgut (Fig. S3D). These observations suggest that elevated expression in this species is due, at least in part, to a novel expression pattern in this digestive organ.

Functional contributions of kraken and Alkbh7 to OA resistance

Given the expression differences observed in the naturally and experimentally evolved OA-resistant drosophilids, we asked whether artificially modulating gene expression would impact OA tolerance. We first performed ubiquitous RNAi of these genes in D. melanogaster (Fig. S4A) and assessed the effects on resistance to OA using a short-term, plate-based assay (Jones 1998) (see Materials and Methods). krakenRNAi led to significantly diminished OA resistance compared to genetic controls, while Alkbh7RNAi did not have a discernible effect (Fig. 5A). We next generated deletion mutants for these genes in D. melanogaster (Fig. S4B). Similar to the RNAi experiments, kraken mutants, but not Alkbh7 mutants, displayed reduced resistance to OA (Fig. 5B). All genotypes remained fully viable when no OA was added (Fig. S5A).

Functional validation of the contribution of kraken and Alkbh7 to OA resistance.

A-C Survival curves of adult flies of the indicated genotypes tested in the plate assay with 2 μl OA for kraken and Alkbh7 RNAi (A), mutants (B) and overexpression (C). Each replicate consisted of 10 2-7 day-old female flies. N replicates ≥6, n flies ≥60. Shading indicates 95% confidence intervals. Genotypes: A act-Gal4/+;UAS-krakenRNAi/+, UAS-krakenRNAi/+, act-Gal4/+, act-Gal4/+;UAS-Alkbh7RNAi/+, UAS-Alkbh7RNAi/+, act-Gal4/+ B kraken1, Alkbh71, Canton-S (wild-type genetic background control for both mutants), C act-Gal4/+;UAS-krakeno/e/+, UAS-krakeno/e/+, act-Gal4/+, act-Gal4/+;UAS-Alkbh7o/e/+, UAS-Alkbh7o/e/+, act-Gal4/+. Significance is based on the resulting mixed effect Cox regression model. NS p>0.05, *p<0.05, **p<0.01, ***p<0.001. The same data for D. sechellia are shown in both plots in C. D Survival curves of wild-type (Dsec07), and kraken (Dsec\krakenRFP) and Alkbh7 (Dsec\Alkbh7RFP) mutant D. sechellia in the plate assay with 30 μl OA. Statistics as in A. E Long-term survivability of the same genotypes as in D (n = 60 per condition) in control conditions (no OA) (left) and food supplemented with 0.9% OA (right). Statistics as in A. Raw data in Data S7.

Loss-of-function analyses might fail to uncover a role for Alkbh7 in whole animals if there are redundant resistance mechanisms absent from S2R+ cells. We therefore took a complementary approach by overexpressing these genes in D. melanogaster. Strikingly, overexpression of Alkbh7 resulted in survivability of D. melanogaster to levels close to those observed for D. sechellia under the same experimental conditions (Fig. 5C, Fig. S4C), indicating that this gene is sufficient to confer high-level OA resistance. By contrast, kraken overexpression did not have any detectable impact on OA resistance (Fig. 5C); a similar lack of effect was seen when overexpressing D. sechellia kraken (Fig. S2E). The absence of a phenotypic effect upon kraken overexpression is surprising, given its elevated expression in D. sechellia and in evolved D. simulans but this could reflect either insufficient overexpression levels (Extended Data Fig.O4c) and/or genotype-by-genotype interactions that are not recapitulated when manipulating a single locus in a D. melanogaster background.

A key question is whether these genes contribute to OA resistance of D. sechellia. Implementing CRISPR/Cas9-based genome-engineering in this species (Auer, et al. 2020; Auer, et al. 2021), we generated and validated null mutants in these genes in D. sechellia (Fig. S4B). These mutant flies were homozygous viable and fertile, with no obvious morphological or behavioural abnormalities, allowing us to test their resistance to OA in the plate assay. Although we did not detect any significant difference to genetically-matched wild-type flies (Fig. 5D), this was anticipated given the polygenic nature of OA resistance in D. sechellia (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015), and the modest or absent phenotypes of equivalent mutants in D. melanogaster (Fig. 5B). Likewise, rearing D. sechellia null mutants on noni pulp (Fig. S5B), we observed no significant differences between mutants and wild-type flies over >30 days, consistent with the conclusion that loss of either gene alone does not compromise D. sechellia’s capacity to persist in its natural host fruit.

We reasoned that more subtle differences might be detected by monitoring D. sechellia survivability at higher OA concentrations over several days. We therefore exposed flies to food containing 0.9% OA, approximately threefold higher than the ∼0.3% OA reported for ripe noni fruit (3.06 g OA/kg fruit) (Pino, et al. 2009). Indeed, we observed higher mortality rates of both kraken and Alkbh7 mutants compared to genetically matched wild-type flies over a period of 6 days (Fig. 5E). By contrast, both mutants survived equally well as controls in the absence of OA over this time period (Fig. 5E). To determine whether the observed OA sensitivity was attributable to a generalised reduction in fitness, we examined survivorship under several conditions of thermal stress (Fig. S5C): mutant and wild-type flies showed no significant differences across all temperatures tested, indicating that the increased mortality observed upon OA exposure does not result from an overall physiological weakness of the knockout lines. These data provide the first evidence for the contribution of specific genes to OA resistance in D. sechellia. However, future allelic swap studies with D. simulans or D. melanogaster would be required to assess if and how they contribute to species-specific differences in OA sensitivity (e.g., through structural and/or regulatory differences in the genes).

Discussion

Identifying specific loci underlying phenotypic evolution is challenging due to the polygenicity of the vast majority of adaptive traits, with recent advocacy of new experimental paradigms to bridge genotype-phenotype relationships (Tautz, et al. 2026). Investigating the evolution of chemical toxicity is an attractive model trait because of the relatively easy phenotypic read-out. To gain insight into the genetic basis of susceptibility and resistance to a natural toxin, OA, we have intersected the results of selection over vastly different timescales: weeks in cell lines, years in experimentally-evolved laboratory strains, and millennia during speciation.

Unlike insecticides that activate or antagonize broadly-expressed neuronal ion channels to cause organismal death, the mode of action of OA appears to be mechanistically very different, capable of inducing death at the level both of whole organisms and of cultured cells. Beyond insect cell lines (our work and (Kaczmarek, et al. 2022)), OA is also toxic to budding yeast (Mota, et al. 2024) and bacteria (Liu, et al. 2013), suggesting that this toxin acts via a more fundamental cellular mechanism. Indeed, a genome-wide, knock-out screen in yeast for genes required for tolerance of OA highlighted contributions of several core cellular processes, such as vesicle trafficking, mitochondrial function or chromatin regulation (Mota, et al. 2024), reminiscent of our CRISPR screen results. Consistent with this idea, the results of our experimental evolution and CRISPR screening, as well as earlier mapping efforts in D. sechellia/D. simulans hybrids (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015) revealed multiple genomic regions/genes contributing to the resistance to OA.

Considered separately, the experimental evolution and cell-based CRISPR screens identified many candidates. That there was very limited overlap between the screen results is unsurprising given their very different designs. The loss-of-function, cell-based screen of course has several limitations, as it cannot identify genes that are not expressed in the cell line, those essential for cell viability in the absence of OA, nor genes that function in biological processes only pertinent at a tissue/organismal level (e.g., cuticle biogenesis, neuronal signalling). Nevertheless, both lists are likely to contain multiple genes relevant to understanding OA susceptibility and resistance, and hence mode(s) of action of OA. Here, we focused on two resistance gene candidates, kraken and Alkbh7, which were common hits of both screens. We note, however, that neither of these genes were identified in previous studies of OA resistance (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015), which could reflect several factors, such as differences in OA toxicity assay conditions, developmental stages, species and molecular readouts (DNA or RNA), and the evident multilayered, polygenic nature of this trait (discussed further in (Marconcini, et al. 2026)).

The (partial) necessity, but insufficiency, of kraken, and lack of necessity but sufficiency of Alkbh7 to confer OA resistance implies fundamentally different roles of these proteins. The expression of kraken in the gut and renal system supports a hypothesis of this protein as a detoxification enzyme, although future work will be necessary to determine whether Kraken has a direct role in OA degradation. The mild loss-of-function phenotype of kraken in both D. melanogaster and D. sechellia likely reflects substantial redundancy of defence mechanisms, including other potential detoxification mechanisms, especially given the plethora of enzymes expressed in the fly gut and Malpighian tubules (Lemaitre and Miguel-Aliaga 2013; Dow, et al. 2022). Indeed, extremely few associations between such enzymes and substrates and/or physiological roles have previously been defined. One exception is the Cytochrome P450 gene Cyp6g1, which has a key role in Malpighian tubules in conferring resistance to DDT (Yang, et al. 2007). Our data on Kraken therefore provide one of the first links between a natural toxic chemical and a putative detoxification enzyme.

The mitochondrial Alkbh7 enzyme likely has a distinct role in OA resistance. Insights from mammals are informative: mice lacking Alkbh7 are obese, with higher body fat levels, proposed to result from a defect in metabolism of fatty acids (Solberg, et al. 2013). Molecularly, one study suggests that Alkbh7’s metabolic role stems from its regulation of RNA editing to control mitochondrial protein translation (Zhang, et al. 2021), although it is unclear whether this is its only function. While the Alkbh7 loss-of-function phenotype is mild in drosophilids, it is striking that artificially elevated expression in D. melanogaster is sufficient to produce D. sechellia-like levels of OA resistance. This result implies a potent effect of Alkbh7 on the mitochondrial metabolic network to modulate toxin resistance, perhaps through control of fatty acid metabolism. While such ideas can be tested in future studies, we note that defects in lipid metabolism and mitochondrial function have been observed upon exposure to other types of insecticide (Martelli, et al. 2020; Martelli, et al. 2022). Intriguingly, our implication of Alkbh7 in the niche adaptation of D. sechellia has a potential parallel in bats, where this gene displays evidence of positive selection during dietary diversification of different species (Gutierrez-Guerrero, et al. 2020).

Taken together, our work has demonstrated the power of orthogonal selection approaches in the laboratory to understand toxin resistance, which might be relevant to understand a historic paradigm of this phenomenon (R’Kha, et al. 1991), as well as offering clues as to how the natural toxin OA exerts its lethal effects. Moreover, the methodology – notably cell-based CRISPR screening, which can now be extended to both gain-of-function approaches (Xia, et al. 2023) and mosquito cell lines (Viswanatha, et al. 2021) – is applicable to explore the molecular mechanisms of susceptibility and resistance of insects to the wide-range of artificial toxic chemicals used in the environment (Gandara, et al. 2024). Such knowledge will have application in the future characterization of novel types of insecticides.

Materials and Methods

Drosophila strains and husbandry

Drosophila stocks were maintained on standard wheat flour–yeast–fruit–juice medium under a 12 h light:12 h dark cycle at 25°C. Wild-type strains, as well as mutant and transgenic lines used in this study are listed in Table S1. Experimental evolution was performed using a subset of published D. simulans isogenic strains (Signor, et al. 2018) (Table S2).

Generation of mutant and transgenic flies

Loss-of-function mutants

we generated D. melanogaster mutants for kraken and Alkbh7 using in vivo CRISPR/Cas9 mutagenesis (Port, et al. 2020) by crossing flies expressing Cas9 in the germline (nanos-Cas9) to those expressing sgRNAs targeting the desired gene. For kraken, we used P{hsFLP}1,y1,w1118;P{HD_CFD02326}attP40 as a source of sgRNAs driven by Act5C-GAL4; for Alkbh7, we used the CG14130 sgRNA line described in (Lenče 2021). Mutants were validated by genomic PCR and sequencing (Fig. S4B).

D. sechellia mutant flies were generated by WellGenetics Inc. using modified methods of (Kondo and Ueda 2013). For kraken (GM16772/LOC6617307) the upstream sequence CTGTAATGCTGGACGCGGAT[TGG] (PAM in square brackets) and the downstream sequence CCTGGTCACCCCGGATCGAG[TGG] were cloned separately into a U6 promoter plasmid. The 3xP3-RFP cassette, which contains a floxed 3xP3-RFP flanked by a 5’ homology arm (999 bp: from -1043 nt to -45 nt relative to the kraken ATG) and a 3’ homology arm (1063 bp: +1388 nt to +2450 nt) were cloned into pUC57-Kan as a donor template for repair. For Alkbh7 (GM25300/LOC6605344) the upstream sequences ACTAATAAGAGTCGACCGGG[TGG], AACAGTCAAACAGCTGTTCC[TGG] and downstream sequences GTCCCAGTTTCAGGGTACGC[TGG], ATGGAAATGCGACGTGTCCT[GGG] were cloned separately into a U6 promoter plasmid. The 3xP3-RFP cassette, which contains a floxed 3xP3-RFP, 5’ homology arm (1037 bp: -1121 nt to -85 nt relative to the Alkbh7 ATG) and 3’ homology arm (964 bp: +940 nt to +1903 nt) were cloned into pUC57-Kan. DNA plasmids encoding the sgRNAs and the corresponding homology-directed repair template were microinjected into embryos of Dsec.07 pBAC(nos:Cas9,3P3-YFP) (Auer, et al. 2020). In both cases, a single F1 line expressing the 3xP3-RFP selection marker was obtained, which was validated by genomic PCR and sequencing (Fig. S4B).

D. melanogaster and D. sechellia mutant flies were backcrossed for six generations to Canton-S and Dsec07, respectively.

Transgenic flies

to make the UAS-Dmel\kraken and UAS-Dsec\kraken transgenes, we amplified (using oligonucleotides listed in Table S6) the regions from the start to the stop codons from w1118and Dsec07 genomic DNA and subcloned them into a pUASTattB vector (Bischof, et al. 2007). Transgenic flies were obtained by phiC31 integrase-mediated transgenesis into attP2 performed by BestGene Inc. and verified by sequencing.

Toxin resistance assays

Tube assay

resistance to OA for the experimental evolution setup was scored adapting a previous protocol (Amlou, et al. 1997; Colson 2004). In brief, flies were lightly anesthetized with CO2 and ∼50 animals were placed in standard cultured tubes (25 mm diameter × 95 mm height, Milian SA) and provided with 1 ml of a 3% glucose solution on a tissue paper (Kimtech, 7552). 3 μl of pure OA (Sigma-Aldrich, CAS 124-07-2, C2875) were directly pipetted onto the tissue while the flies remained anesthetized. Direct contact with the acid was scrupulously avoided until all flies recovered. The fraction of alive flies was scored after 24 h. For the experimental evolution experiment, the surviving flies were collected (see below).

Plate assay

to measure OA resistance in the genetically-manipulated D. melanogaster and D. sechellia strains, we used a plate assay (Jones 1998). Compared to the experiments in tubes, this assay offers greater temporal resolution and higher dynamic range (in contrast to the experimental evolution experiment, we did not strive to maintain a certain fraction of surviving flies), as well as greater statistical power from tests of fewer animals. Ten female flies were lightly anesthetized with CO2 and placed in a Petri dish (60 mm diameter × 15 mm height) in which 2 μl of pure OA (or 30 μl for D. sechellia) were spotted in the centre of the underside of the lid. The flies were allowed to recover (typically <5 min), and immobile/dead were scored every 5 min for 1 h. The survival curves were calculated and plotted using the R package survminer (10.32614/CRAN.package.survminer). The statistical analysis was conducted fitting a mixed effect Cox regression model as implemented in the R package coxme (Therneau 2024). Genotype was included as a fixed effect, and replicate dishes were modelled as a random effect.

Noni resistance assay

noni fruit was collected from M. citrifolia plants grown in the University of Lausanne greenhouses. Five ripe (pale yellow) fruits were homogenized together, and the pulp puree frozen at -20°C. The noni assay was performed with the same setup as for tube assay described above, using 2 g of thawed noni puree placed on the bottom of the tube. Noni contains 3.06 g OA/kg fruit (Pino, et al. 2009); assuming an equivalent quantity in our fruits, 2 g noni pulp should contain 6.73 µl OA, approximately double the amount used for the experimental evolution between generations 1 to 25, and two thirds of the concentration during generations 30-50.

Long-term assay

To assess the long-term effects (i.e., beyond 24 h) of OA or noni pulp, two-day-old mated female flies were individually housed in vials containing either food supplemented with OA or 2g of noni puree. OA-supplemented food was prepared by thoroughly mixing OA into Formula 4-24® Instant Drosophila Medium, Blue (Carolina Biological Supply) to a final concentration of 0.9%. Flies were flipped into fresh food every two days, and mortality was monitored daily. Statistical analysis was run as for the plate assay.

Fecundity and developmental viability

10 mated adult female flies (4-7 days old) were allowed to lay eggs for 24 h on agar plates (2% agar, 2.5% sucrose in a 1:3 grape juice:water mix). The total number of eggs on each plate was counted. From each plate, 100 eggs were collected and transferred to standard fly food. Emerging adults were counted and removed daily until no further flies emerged.

Lifespan

Two-day-old mated female flies were individually housed in vials containing standard fly food and monitored daily until death. Fresh food was provided every 2–3 days.

Experimental evolution

Selection procedure

to generate the base populations for experimental evolution, we used 93 isogenic lines (Table S2) of the D. simulans panel established by (Signor, et al. 2018). We first crossed each of the lines together following a round-robin design to ensure equal contribution of each line to the starting genetic pool; 5 males and 5 females from each cross were then combined in ten independent replicates (Fig. S1D). These ten populations were maintained in bottles (57 mm diameter ×103 mm height, Milian SA) of standard fly food at high population size (∼1000 individuals per population) to minimize the effect of genetic drift. Each population was subdivided into smaller groups of ∼250 individuals per bottle (i.e., four bottles per population) to reduce competition. At each generation, individuals from the four bottles were mixed to prevent bottle-specific genetic drift. The experimental evolution experiment was initiated 10 generations after the creation of these populations.

From each population, 500 individuals (2-7 day old) were randomly collected and further split into 10 groups of 50 mixed-sex flies (i.e., ∼5000 flies total). Each group of 50 flies was then placed under selection with 3 μl of OA for 24 h (as described above); the survivors were collected and allowed to breed in food vials for 6-8 days, transferring them to fresh vials every 2-3 days. In the next generation, 500 flies per population were collected only from the last seeded vial, to maximize the probability of these flies being parented by the OA-resistant males and females and not offspring generated by females that mated before the selection procedure. These flies were then subjected to the same selection procedure. This process was repeated for 25 generations, and survivability was scored at each generation.

A second round of experimental evolution was performed by increasing the amount of OA to 9 μl between generations 31 and 50; this amount was determined empirically during generation 26-31 to re-establish the survivability close to the initial level (Fig. 1G). All successive generations were maintained following the same procedure.

As a control, the ten original populations were maintained in parallel in the same conditions as previously described (i.e. ∼1000 individuals per population subdivided into smaller groups of ∼250 individuals per bottle).

DNA sequencing and analysis

we pooled ∼200 individual flies per population per time point (generations 0, 25 and 50). Genomic DNA (and total RNA for qPCR, as described below) extraction was performed using the Quick DNA/RNA Miniprep plus kit (Zymo Research, D7003) following the manufacturer’s instructions. Samples were stored at -80°C.

We performed paired-end whole genome sequencing on an Illumina NovaSeq 6000 following Nextera DNA flex preparation, at an average coverage of >200×. Read quality was assessed with Fastqc (Wood 2011) and adapters were trimmed with Trimmomatic (Bolger, et al. 2014) providing the Nextera adapters sequences. We aligned the trimmed reads to the D. simulans reference genome (NCBI GCF_016746395.2) using bwa-mem (Li 2013) and the results were sorted and compressed using samtools (Danecek, et al. 2021). Duplicate reads were then removed with Picard MarkDuplicates (https://broadinstitute.github.io/picard/). Unplaced scaffolds and alignments with mapping quality <20 were removed with samtools. We then combined all the aligned sequences into a samtools pileup file and split this into individual chromosomes. Regions with zero coverage were removed with custom scripts. Sync files for each chromosome were generated with the mpileup2sync.pl script implemented in Popoolation2 (Kofler, et al. 2011) and used as input files for all the Evolve-and-Resequence analyses.

To explore the genetic variation of our samples we generated vcf files with vcftools 0.1.17 (Danecek, et al. 2011) and binary files with plink1.9 (Purcell, et al. 2007; Purcell 2020). We calculated the identity-by-state (IBS) matrix as implemented in plink1.9 (function -distance ‘ibs’).

Evolve-and-Resequence analysis

allele frequency changes between generation 0 and generations 25 or 50 were investigated via Popoolation2 and Bait-ER (Barata, et al. 2023).

We performed a Cochran–Mantel–Haenszel (CMH) test via the cmh-test.pl script implemented in Popoolation2 comparing base and evolved populations with parameters --min-count 12 and --min-coverage 50. We estimated a plausible significance threshold approximating the top 0.000005% of the simulated p-values based on the effect of genetic drift alone by running genome-wide forward simulations under neutrality in mimicree2 (Vlachos and Kofler 2018). We used the same genomic information from the original 93 lines as input and the average real survivability ratios observed throughout the selection experiment as –selection-regime parameter and 500 as –population-size parameter (Fig. S1B).

For Bait-ER, the input sync files were filtered to match --min-count 12 and --min-coverage 50 and scaled to a uniform coverage of 100 as in (Vlachos, et al. 2019) to avoid numerical problems due to high coverage. We ran the algorithm on each chromosome separately. Bait-ER models the effect of genetic drift directly using estimates of the effective population size (Ne). We calculated Ne estimates for each chromosome in each replicate population with the R package poolSeq (Taus, et al. 2017) and used the median value of each chromosome as “Population_size” parameter (Table S3). Other parameters included “Sampling correction” = 0 (i.e., binomial) and “Prior parameters” =0.001 (shape), 0.001 (rate). Data points were considered significant when the absolute natural logarithm of the Bayes Factor exceeded the conservative threshold of log(0.99/0.01) as in (Barata, et al. 2023).

The linkage disequilibrium decay in the 10 evolved populations was analysed across the whole genome using PopLDdecay (Zhang, et al. 2019). Linkage disequilibrium maps of focal regions were generated with LDBlockShow (Dong, et al. 2021) using the Solid Spine method based on D’ statistics originally introduced in Haploview (Barrett, et al. 2005).

For the genome-wide screen, all the genes within selection blocks were considered potential candidates. Based on the results on linkage disequilibrium decay, a block was arbitrarily defined as a genomic interval where at least two significant SNPs were found at a distance less than 50 kb. Consecutive SNPs respecting this definition were grouped into larger blocks and visualized as circus plots using the R package circlize (Gu, et al. 2014). The genes overlapping such blocks were extracted using BEDtools intersect (Quinlan and Hall 2010). Orthologous genes in D. melanogaster were retrieved using a combination of the OMA browser (Altenhoff, et al. 2021), DAVID (Huang da, et al. 2009; Sherman, et al. 2022), BLAST (Camacho, et al. 2009), NCBI (Sayers, et al. 2025) and FlyBase (Ozturk-Colak, et al. 2024), and subsequent manual curation.

Genome-wide CRISPR/Cas9 screen in D. melanogaster S2R+ cells

Library design, construction, and transfection

design, synthesis and cloning of the genome-wide sgRNA library are previously described (Viswanatha, et al. 2025). To determine sub-lethal doses of OA appropriate for screening, the effect of a gradient of OA supplementation into media on attP+ Cas9+ cells was determined using CellTiter-Glo (Promega) following the manufacturer’s protocol; luminescence measurements were made using a Spectramax Paradigm (Molecular Devices).

For the screen, attP+ Cas9+ D. melanogaster S2R+ cells (RRID:CVCL_UD30) were transfected using Effectene and a 1:1 molar ratio of sgRNA library and pIntAC, as described (Viswanatha, et al. 2025). Briefly, transfected cells were distributed in 5 ml aliquots across ten 100 mm culture dishes and incubated overnight, then 5 ml of fresh media was added to each plate. After 4Odays, cells were subjected to puromycin selection for 12 days, subculturing every 4Odays, to establish a stable sgRNA-expressing cell library. For the selection assay, 2.5O×O107 cells were plated onto four 15 cm diameter cell culture dishes to ensure sufficient sgRNA coverage, with each sgRNA represented approximately 1000 times (Viswanatha, et al. 2025).

OA selection assay

library-containing cells were left untreated or treated with either 0.075 μg/μl or 0.05 μg/μl OA (Sigma-Aldrich, CAS 124-07-2, C2875). Each replicate consisted of four 15 cm diameter cell culture dishes seeded with cells, achieving ∼1000× coverage. Two replicate sets were included for each of the two OA treatment conditions and one replicate for the untreated control. During the selection period, cells were split 1-2 times per week to avoid overgrowth. An aliquot of cells from each replicate was collected and stored for genomic prep following each split and at the end of the assay. The final samples were collected after 26 days of OA or control treatment.

Genomic DNA extraction and sequencing

genomic DNA was extracted from samples collected during the screen and at the final endpoint using a Zymo Quick-gDNA Miniprep kit (Zymo Research, D3025). DNA fragments containing the sgRNA sequences were amplified by PCR using what we calculated to be at least 1000 genomes per sgRNA as templates for each sample. PCR fragments were in-line barcoded using a previously described approach (Viswanatha, et al. 2018; Viswanatha, et al. 2019; Viswanatha, et al. 2025). Next-generation sequencing (Illumina NextSeq) was performed at the Biopolymers Facility at Harvard Medical School (RRID:SCR_007175). FASTQ sequence files were demultiplexed using in-line barcodes as sequence tags as described previously (Viswanatha, et al. 2018; Viswanatha, et al. 2025). All further processing was done using MAGeCK (v.0.5.4 and 0.5.9.4) (Li, et al. 2014) to generate readcount files and perform robust rank aggregation (RRA) analysis and Maximum likelihood estimation directly on the generated readcount files using default parameters. Plasmid readcounts were used to normalize RRA values.

After removing essential genes at a 2% FDR (2,517 genes; Data S4), we considered as candidates the top 5% genes overlapping across the two conditions that were over-represented or under-represented in the OA-exposed cell populations compared to the control cells (i.e., representing “susceptibility” and “resistance” genes, respectively). The gene-set enrichment analysis was performed with PANGEA 1.1 Beta using the Drosophila GO subsets (GO slim) and a p-value cut-off of 0.05 (Hu, et al. 2023).

qPCR analysis

Reverse transcription of total RNA (extracted as described above) was performed with the PrimeScript first-strand cDNA synthesis kit (Takara, 6110A) following the manufacturer’s instructions. Samples were stored at -80°C.

Quantitative PCR experiments were conducted on a QuantStudio 6 Flex Real-Time PCR System using the PowerUp™ SYBR™ Green Master Mix following the manufacturer’s instructions. Each sample consisted of at least three biological replicates. We performed three technical replicates per reaction and removed replicates where the standard deviation was equal or greater than 0.5. We determined target specific amplification efficiencies for each gene (Table S4, Table S6). Relative quantification and statistical analysis were performed with the software qbase+ v3.4 (Hellemans, et al. 2007) using Calibrated Normalized Relative Quantities (CNRQs). Normalization was performed on the geometric mean of the relative quantities for the three most commonly-used reference targets: Act5C, αTub84B and RpL32 (Lu, et al. 2018). We assessed reference target stability by the geNorm expression stability value (M) and the coefficient of variation of the normalized reference gene relative quantities (CV) (Table S5) (Vandesompele, et al. 2002; Hellemans, et al. 2007).

Population genetics analysis

Publicly available WGS data for 41 D. sechellia and 20 D. simulans genomes were downloaded from the Sequence Read Archive (PRJNA215932 and PRJNA395473) (Rogers, et al. 2014; Schrider, et al. 2018) with fasterq-dump (Leinonen, et al. 2010). FASTQ files were processed and aligned as described in the “Experimental evolution” section. Aligned files were phased with samtools phase (Danecek, et al. 2021) and the genes of interest were extracted with pilon (Walker, et al. 2014) to find variation among populations. CDS sequences were aligned based on amino acid information using translatorX (Abascal, et al. 2010). Gene sequences were aligned using Clustal Omega (Madeira, et al. 2024). McDonald-Kreitman test, Tajima’s D and Fu and Li’s statistics were calculated in DNAsp 6 (Rozas, et al. 2017). FUBAR analyses (Fast, Unconstrained Bayesian AppRoximation) were performed in HyPhy 2.5 (Kosakovsky Pond, et al. 2020). The consensus sequences of D. simulans and D. sechellia were computed by SnapGene (www.snapgene.com) “export consensus” with a >50% threshold. The sequence alignment and secondary structures were depicted using ESPript 3 (Robert and Gouet 2014).

RNA fluorescence in situ hybridization

To detect kraken mRNA in situ, we used the RNAscope Multiplex Fluorescence Detection Reagents v2 (Advanced Cell Diagnostics, 323110) and RNAscope H2O2 and Protease Reagents (Advanced Cell Diagnostics, 322381). Surfaces and equipment were cleaned with RNAzap (Invitrogen, AM9780) prior to dissection. Intestinal tissues were dissected from D. melanogaster (Canton S) and D. sechellia (D. sechellia 07) females in RNAlater (Sigma-Aldrich, R0901) diluted 1:50 in DEPC-treated water and briefly rinsed in 1× PBS (prepared in ultrapure water) (ThermoFisher Scientific, 70011044). Dissected tissues were mounted on poly-L-lysine-coated slides in 1× PBS and fixed immediately in cold 4% formaldehyde (Sigma-Aldrich, F8775) for 30 min. The fixative was removed, and tissues were incubated with few drops of hydrogen peroxide for 10 min at room temperature (RT). Tissues were then washed 3× 5 min in wash buffer (Advanced Cell Diagnostics, 320058) preheated to 40°C. For tissue dehydration, slides were incubated with 200 μl of ethanol solutions in the following order: 50% ethanol for 5 min at RT, 70% ethanol for 5 min at RT and 100% ethanol for 5 min at RT (repeated twice). Slides were air-dried for 5 min, and samples incubated with few drops of Protease III for 30 min at RT.

To hybridise probes, tissues were washed in 1× PBS for 3× 5 min. 50 μl of C1 (D. melanogaster-kraken-RA (NM_057917), Advanced Cell Diagnostics, 1850601-C1); and 1 μl of C2 (D. melanogaster-Rpl32 (NM_079843.4), Advance Cell Diagnostics, 533451-C2) RNAscope FISH probes were preheated to 40°C for 10 min, cooled at RT for 5 min, applied to tissues and the slides were placed in a humidified chamber for hybridization at 40°C for 2 h. Slides were hybridized with AMP reagents in sequential steps. A few drops of AMP1, warmed to RT, were applied to slides, which were incubated at 40°C for 30 min, followed by washes in wash buffer 3× 5 min. Next, a few drops of AMP2, warmed to RT, were applied to slides and incubated at 40°C for 30 min, followed by washes in wash buffer 3× 5 min. Finally, a few drops of AMP3, warmed to RT, were applied to slides, and incubated at 40°C for 15 min, followed by washes in wash buffer 3× 5 min.

A few drops of HRP-conjugated probe was applied for C1 and slides were incubated at 40°C for 15 min. After, rinsing in wash buffer 3× 5 min, Opal dyes 520 nm (Akoya Biosciences, OP-001001) and/or 570nm (Akoya Biosciences, OP-001003 (1:1000 in TSA multiplex buffer, Advanced Cell Diagnostics, 322809) were applied to slides and incubated at 40°C for 30 min. Slides were then washed and treated with HRP blocker (a few drops per slide) for 15 min at 40°C and washed in wash buffer for 3× 5 min. Tissues were counterstained with DAPI for 5 min at RT and washed once in the wash buffer, before mounting in Vectashield (Vector Laboratories, H1000).

A Leica STELLARIS-DIVE confocal microscope was used to generate all confocal images, using a water-immersive 40× objective or an oil-immersive 63× objective. The images were acquired using both HyDs as well as PMTs tailored for the fluorophores of each sample.

Statistics and reproducibility

Data were analysed and plotted in R 4.0.3 (R Core Team 2021).

eLife Assessment

This valuable study advances our understanding of genes contributing to Drosophila resistance to octanoic acid, a primary toxin present in Morinda fruit, which is the natural host plant for Drosophila sechellia, a species that has become a model for understanding evolutionary specialization. The authors provide solid results from an original combination of experimental evolution and cell-based CRISPR screens. This work will be of interest to the Drosophila community and researchers interested in the genetic basis of polygenic traits.

Reviewer #1 (Public review):

Marconcini et al. report results of an ambitious study on the genetic mechanisms that contribute to resistance of Drosophila flies to the toxin octanoic acid (OA). This study was motivated by two observations: first, Drosophila sechellia, a close relative of D. melanogaster, has evolved specialized feeding on fruits of Morinda citrifolia, which contain high concentrations of OA and second, that artificial selection on Drosophila simulans, a sister species of D. melanogaster, can generate higher resistance to OA. Previous studies had performed genetic mapping studies between D. simulans and D. sechellia that implicated certain genomic regions in resistance to OA and, in particular, implicated several Osiris gene paralogs as contributing to resistance, though the molecular mechanisms of resistance remain unclear. In this study, Marconcini et al. performed two major experiments. First, they performed evolution-and-resequence on Drosophila simulans populations exposed to OA for 50 generations and identified candidate regions with excessive shifts in allele frequencies as candidate regions containing OA resistance genes in D. simulans. Second, they performed a CRISPR knock-out screen in a D. melanogaster cell line to identify genes that contribute to OA resistance and susceptibility.

Evolve-and-resequence yielded many candidate genomic regions with extreme allele frequency shifts, which may be regions containing OA resistance genes, or linked genes, or regions that happen to show a strong shift in all replicate populations by chance. As the authors note, detecting significant shifts in allele frequencies is a challenging problem, and the authors use two measures of allele frequency shifts (the Cochran-Mantel-Haenszel method and Bait-ER) and perform simulations under neutrality to estimate a reasonable significance threshold. I am not entirely convinced by this method of estimating significance levels, because the simulations involve assumptions that may not be met by the real populations. I would think that a permutation test would provide an assumption-free method of estimating significance levels. I have tried to think whether there is something about the design of these experiments that would preclude the use of permutation tests (which are used widely for genome-wide studies, such as QTL), but I can't think of one. Perhaps the authors are aware of a reason permutation tests would be invalid here, and if so, they should state this reason.

There is overlap between regions detected by the two methods, but the methods disagree for many regions. The authors state that a "majority of prominent peaks were found by both methods," but I am unclear on what "prominent" means here. It would be more helpful to be more quantitative about the extent of overlap.

The authors hypothesized that the response would be at similar genomic loci in all populations (line 222). It seems at least possible that epistatic interactions would lead to different combinations of alleles evolving in each population. I wonder if it would be possible to test whether there is heterogeneity in the responses across the replicate populations.

The evolve-and-resequence method yielded many possible regions contributing to OA resistance in D. simulans, but perhaps too many regions to test directly or even to build sensible hypotheses about the genes involved. Thus, the authors performed a second experiment to try to narrow down the list of possible candidate genes. They performed a CRISPR knockout screen in a D. melanogaster cell line for genes that contribute to resistance or susceptibility to OA. The authors identify several limitations of this experiment, but they nonetheless identified several genes where knockouts contribute to OA susceptibility or resistance. Intersecting top hits with regions that experienced selection identified two "resistance" genes: kraken and Alkbh7. The selection hit at kraken is quite compelling, whereas the evidence at Alkbh7 is less strong because only two SNPs were marginally significant. Further functional assays, including gene knockouts in D. melanogaster and D. sechellia, provide some support for the claim that both of these genes can contribute to resistance to OA in flies.

Beyond the few issues raised above, I do not have significant questions about methodology or the results. I do think, however, that the authors should be more conservative about the implications and significance of their results. For example, on line 139, the authors claim that this intersection approach provides a "powerful paradigm to investigate ecotoxicology." I am not sure I agree that the identification of two genes that may contribute to OA resistance, after a seemingly heroic selection experiment and CRISPR screen, suggests that this method is all that powerful. It seems that most of the genes that contribute to the selection response remain unidentified.

Finally, given that one motivation of this project was to identify genes that contribute to evolved resistance to OA, I am surprised that the authors did not generate CRISPR alleles of kraken and Alkbh7 in D. simulans and then use these together with the existing alleles in D. sechellia to perform reciprocal hemizygosity tests to determine if these two genes actually contribute to evolved resistance in D. sechellia. This test is simpler to perform and may be more sensitive than the allelic replacement that the authors propose (lines 446-449).

Reviewer #2 (Public review):

Summary:

The authors studied the resistance against octanoic acid, a compound of the noni fruit, in D. simulans, using experimental evolution and resistance/susceptibility in D. melanogaster cells. They identified novel candidate genes and performed functional tests.

Strengths:

The idea of using experimental evolution of a non-resistant species to develop resistance is interesting, and the idea of narrowing down a large list of candidate loci by CRISPR-based gene knockout in cell culture is innovative. The reviewer also liked the (easy) follow-up experiments to validate the results.

Weaknesses:

The reviewer is not convinced of the conceptual idea behind their approach: the intersection of the two approaches implicitly assumes that null alleles (or at least compromised alleles) should be selected during experimental evolution. The reviewer considers this unlikely, and the authors made no attempt to test this implicit hypothesis in their data. Along the same lines, it is not clear how to reconcile an upregulation of candidate genes in resistant flies with the knockout experiments.

The experiments to validate the effect of candidate genes did not match the experimental evolution conditions.

The statistical analysis suffers from some problems and an insufficient description of the analyses performed.

Although D. simulans GWAS data are available, the authors did not make an attempt to estimate the effect of selected variants in the candidate genes in the GWAS data set.

The reviewer would have liked to see more connection between the experimental evolution and the GWAS data. As some D. simulans genotypes have similar resistance to D. sechellia, it would have been interesting to test whether this genotype contributed to the observed resistance.

At several places, the authors discuss the challenge of studying a polygenic trait, but at the same time, they claim to have detected and validated candidate genes. It would be helpful if the authors could discuss why they consider that their assays could really detect the contribution of single loci to the polygenic trait. In particular, when GWAS did not detect their candidate genes.

It is not clear to the reviewer why the authors did not pay more attention to the highly significant peaks emerging from the experimental evolution study. Their functional validation would have been biologically more plausible.

Impact:

Given the obvious challenges of functional testing of polygenic traits and the clear limitations of the interpretation of the results, the study will be helpful for future studies aiming to characterize polygenic traits. Unfortunately, the results are just another piece of controversial results regarding resistance against octanoic acid, a trait that is rather easy to evaluate.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Marconcini et al. report results of an ambitious study on the genetic mechanisms that contribute to resistance of Drosophila flies to the toxin octanoic acid (OA). This study was motivated by two observations: first, Drosophila sechellia, a close relative of D. melanogaster, has evolved specialized feeding on fruits of Morinda citrifolia, which contain high concentrations of OA and second, that artificial selection on Drosophila simulans, a sister species of D. melanogaster, can generate higher resistance to OA. Previous studies had performed genetic mapping studies between D. simulans and D. sechellia that implicated certain genomic regions in resistance to OA and, in particular, implicated several Osiris gene paralogs as contributing to resistance, though the molecular mechanisms of resistance remain unclear. In this study, Marconcini et al. performed two major experiments. First, they performed evolution-and-resequence on Drosophila simulans populations exposed to OA for 50 generations and identified candidate regions with excessive shifts in allele frequencies as candidate regions containing OA resistance genes in D. simulans. Second, they performed a CRISPR knock-out screen in a D. melanogaster cell line to identify genes that contribute to OA resistance and susceptibility.

Evolve-and-resequence yielded many candidate genomic regions with extreme allele frequency shifts, which may be regions containing OA resistance genes, or linked genes, or regions that happen to show a strong shift in all replicate populations by chance. As the authors note, detecting significant shifts in allele frequencies is a challenging problem, and the authors use two measures of allele frequency shifts (the Cochran-Mantel-Haenszel method and Bait-ER) and perform simulations under neutrality to estimate a reasonable significance threshold. I am not entirely convinced by this method of estimating significance levels, because the simulations involve assumptions that may not be met by the real populations. I would think that a permutation test would provide an assumption-free method of estimating significance levels. I have tried to think whether there is something about the design of these experiments that would preclude the use of permutation tests (which are used widely for genome-wide studies, such as QTL), but I can't think of one. Perhaps the authors are aware of a reason permutation tests would be invalid here, and if so, they should state this reason.

Significance thresholds have been estimated using a variety of approaches in the literature, including false discovery rate (FDR) control, simulations under neutral drift with predefined cut-offs, arbitrary significance thresholds, haplotype-based analyses, permutation tests, amongst other methods. Our choice was guided by a review (doi:10.1186/s13059-019-1770-8), which compared the performance of several of these approaches and found that the relatively simple assumptions underlying the Cochran–Mantel–Haenszel (CMH) test often performed as well as, or better than, more complex alternatives. As with most significance thresholds used in genome-wide analyses, the CMH threshold applied here is ultimately based on a degree of arbitrariness, although we aimed to be as stringent as possible. As a complementary method, we used Bait-ER; here the threshold used followed the recommendations provided in the original publication describing this method (doi:10.1111/jeb.14134).

One reason that permutation-based approaches may be less widely adopted than theoretical null models is the extensive linkage disequilibrium among SNPs, which results in large blocks of correlated variants that cannot be considered independent observations (and therefore not shuffled). In principle, permutations could be performed at the haplotype-block level; however, defining haplotype blocks itself requires selecting thresholds or criteria that are often user defined or based on significance cut-offs. Although several tools are available for this purpose, our experience has been that the resulting block definitions remain sensitive to these choices and therefore introduce a comparable degree of arbitrariness. In reality, there is not yet a standard method in the field that has emerged as the “best” one.

There is overlap between regions detected by the two methods, but the methods disagree for many regions. The authors state that a "majority of prominent peaks were found by both methods," but I am unclear on what "prominent" means here. It would be more helpful to be more quantitative about the extent of overlap.

To quantify the agreement between the two methods, we calculated the overlap in genomic coverage (base pairs) between CMH and Baiter candidate regions. For G25, the two methods shared 1.90 Mb of candidate regions. The overlap encompassed 65.3% of the genomic span identified by CMH and 64.1% of the span identified by Bait-ER. For G50, the methods shared 4.77 Mb of candidate regions. The overlap encompassed 63.6% of the CMH candidate span and 99.9% of the Bait-ER candidate span. We have now replaced the admittedly qualitative statement the reviewer highlighted ("majority of prominent peaks were found by both methods”) with this information.

The authors hypothesized that the response would be at similar genomic loci in all populations (line 222). It seems at least possible that epistatic interactions would lead to different combinations of alleles evolving in each population. I wonder if it would be possible to test whether there is heterogeneity in the responses across the replicate populations.

We agree that epistatic interactions could lead to different allelic combinations being favored in different replicate populations. However, both BaitER and the CMH test are designed to detect parallel evolutionary responses across replicates, which was the focus of our study. One approach for testing population-specific responses is the LRT-2 test (doi:10.1534/genetics.118.301824). We did not pursue this analysis because it would likely generate many additional candidate loci, making interpretation more challenging, while providing limited additional insight into the repeatable genomic responses that were the primary focus of this work.

The evolve-and-resequence method yielded many possible regions contributing to OA resistance in D. simulans, but perhaps too many regions to test directly or even to build sensible hypotheses about the genes involved. Thus, the authors performed a second experiment to try to narrow down the list of possible candidate genes. They performed a CRISPR knockout screen in a D. melanogaster cell line for genes that contribute to resistance or susceptibility to OA. The authors identify several limitations of this experiment, but they nonetheless identified several genes where knockouts contribute to OA susceptibility or resistance. Intersecting top hits with regions that experienced selection identified two "resistance" genes: kraken and Alkbh7. The selection hit at kraken is quite compelling, whereas the evidence at Alkbh7 is less strong because only two SNPs were marginally significant. Further functional assays, including gene knockouts in D. melanogaster and D. sechellia, provide some support for the claim that both of these genes can contribute to resistance to OA in flies.

Beyond the few issues raised above, I do not have significant questions about methodology or the results. I do think, however, that the authors should be more conservative about the implications and significance of their results. For example, on line 139, the authors claim that this intersection approach provides a "powerful paradigm to investigate ecotoxicology." I am not sure I agree that the identification of two genes that may contribute to OA resistance, after a seemingly heroic selection experiment and CRISPR screen, suggests that this method is all that powerful. It seems that most of the genes that contribute to the selection response remain unidentified.

We agree with the reviewer and have modified the sentence accordingly. While we believe that integrating the approaches discussed in this paper can provide valuable insights into the genetic basis of ecotoxicological traits, these approaches are not a panacea for traits with highly complex genetic architectures, such as OA resistance. Nevertheless, the identification of two candidate genes with some evidence of contributing to the trait represents a meaningful advance toward understanding its underlying genetic basis.

Finally, given that one motivation of this project was to identify genes that contribute to evolved resistance to OA, I am surprised that the authors did not generate CRISPR alleles of kraken and Alkbh7 in D. simulans and then use these together with the existing alleles in D. sechellia to perform reciprocal hemizygosity tests to determine if these two genes actually contribute to evolved resistance in D. sechellia. This test is simpler to perform and may be more sensitive than the allelic replacement that the authors propose (lines 446-449).

While generating null alleles for the candidate genes in D. simulans is beyond the scope of this revision, we note that we did attempt reciprocal hemizygosity tests using the mutants available in D. melanogaster. However, the resulting hybrids were recovered in low numbers, and these animals were rather weak, making them unsuitable for the severe OA exposure conditions employed in our assays. We agree that, where feasible, future studies should incorporate reciprocal hemizygosity tests, as they represent a powerful approach for validating the contribution of candidate genes to the trait of interest. We have revised the final sentence of the Results section accordingly.

Reviewer #2 (Public review):

Summary:

The authors studied the resistance against octanoic acid, a compound of the noni fruit, in D. simulans, using experimental evolution and resistance/susceptibility in D. melanogaster cells. They identified novel candidate genes and performed functional tests.

Strengths:

The idea of using experimental evolution of a non-resistant species to develop resistance is interesting, and the idea of narrowing down a large list of candidate loci by CRISPR-based gene knockout in cell culture is innovative. The reviewer also liked the (easy) follow-up experiments to validate the results.

Weaknesses:

The reviewer is not convinced of the conceptual idea behind their approach: the intersection of the two approaches implicitly assumes that null alleles (or at least compromised alleles) should be selected during experimental evolution. The reviewer considers this unlikely, and the authors made no attempt to test this implicit hypothesis in their data.

We respectfully disagree with the reviewer’s interpretation of the conceptual idea behind our approach. Our strategy did not assume that experimental evolution selects for null alleles, but rather for any type of variant that could contribute to the trait being selected for (i.e., increases in OA resistance), pointing to candidate genes contributing to the trait. Like many evolve-and-resequence experiments, this approach identified hundreds of candidate genes. This is why we took an orthogonal, genome-wide CRISPR screening approach, where loss-of-function mutations could lead to increases or decreases in OA tolerance of cultured cells. However, we stress that the naturally selected alleles – which could be gain or loss of function – may have much subtler phenotypic effects than the null alleles used for functional validation.

Along the same lines, it is not clear how to reconcile an upregulation of candidate genes in resistant flies with the knockout experiments.

We respectfully disagree that these findings are difficult to reconcile. The observed upregulation of the candidate genes in the selected, resistant D. simulans is consistent with a role in promoting OA resistance, while the knockout experiments (whether in cultured cells or in whole animals) demonstrate that loss of gene function reduces resistance. These observations are complementary: increased expression is associated with enhanced resistance, whereas complete loss of function impairs it. The knockout experiments of kraken and Alkbh7 were intended as functional validation of gene involvement and do not imply that the alleles selected during experimental evolution are loss-of-function alleles (we rather hypothesize that the selected alleles are gain-of-function through some, as yet undetermined, mechanism).

The experiments to validate the effect of candidate genes did not match the experimental evolution conditions.

This is correct. As our results suggest that the phenotype is shaped by multiple genes, such that the effect of any individual gene is likely modest compared to their combined contribution. Consequently, detecting and validating the effect of a single gene requires more stringent OA conditions than those needed to observe the overall phenotypic response over the course of several generations. In addition, the shorter-term plate assay was more practical for higher temporal resolution of the analysis of mortality in the presence of OA.

The statistical analysis suffers from some problems and an insufficient description of the analyses performed.

Although D. simulans GWAS data are available, the authors did not make an attempt to estimate the effect of selected variants in the candidate genes in the GWAS data set.

We agree that comparing the experimental evolution and GWAS results (from our previous work, doi:10.1093/g3journal/jkag032) is of considerable interest. We have now expanded the Discussion to explicitly discuss the relationship between the two datasets. Overall, the overlap between the approaches was limited, although two GWAS candidate genes, bez and CG13003, fall within genomic regions exhibiting significant CMH signals at generation 25 (but not generation 50) of the evolve-and-resequence experiment. (We note that these genes could not have been identified in our CRISPR screen, as they are not expressed in S2R+ cells). More broadly, differences between the GWAS and evolve-and-resequence results likely reflect the distinct evolutionary processes captured by each approach: GWAS maps standing phenotypic variation among isofemale lines, whereas experimental evolution tracks allele frequency changes under sustained selection. Understanding why some signals are shared whereas others are not – whether due to effect size, genetic background, epistasis, pleiotropic costs, or the contribution of initially rare variants – remain important open questions.

The reviewer would have liked to see more connection between the experimental evolution and the GWAS data. As some D. simulans genotypes have similar resistance to D. sechellia, it would have been interesting to test whether this genotype contributed to the observed resistance.

While D. simulans genotypes displayed a range of OA resistance levels, none approached D. sechellia levels of resistance (see Figure 2e from our previous work, doi:10.1093/g3journal/jkag032) (The reviewer might have conflated the data from our GWAS of D. melanogaster strain, shown in Figure 2b of that paper, where some lines of that species exhibit comparable resistance to D. sechellia under the conditions of that assay). Regardless, our experimental evolution data do not provide sufficient resolution to identify the specific favorable alleles underlying the response to selection. Instead, we detect genomic regions containing many linked variants whose frequencies change under selection. Consequently, a direct comparison between evolved genotypes and GWAS-associated genotypes is currently difficult. We note, however, that expression of the D. sechellia kraken allele in D. melanogaster did not produce a significant effect on resistance, suggesting that even the most promising candidate alleles might not have strong effects in isolation.

At several places, the authors discuss the challenge of studying a polygenic trait, but at the same time, they claim to have detected and validated candidate genes. It would be helpful if the authors could discuss why they consider that their assays could really detect the contribution of single loci to the polygenic trait. In particular, when GWAS did not detect their candidate genes.

Our results do not imply that kraken and Alkbh7 are major-effect loci or that they explain a substantial proportion of the phenotypic variation. Rather, our data indicate that these genes make measurable contributions to OA resistance, consistent with the expectation that complex traits are influenced by many loci of individually modest effect. The absence of these genes among the top GWAS candidates does not preclude their involvement, as GWAS and experimental evolution interrogate different aspects of the genetic architecture and differ in their power to detect loci of varying effect sizes and allele frequencies.

It is not clear to the reviewer why the authors did not pay more attention to the highly significant peaks emerging from the experimental evolution study. Their functional validation would have been biologically more plausible.

We agree that the significant peaks identified in the evolve-andre sequence experiment represent promising targets for future investigation. However, these peaks typically span large genomic regions containing tens to hundreds of genes, making it difficult to prioritize individual candidates based on the experimental evolution data alone. In this work, we focused our functional validation on genes independently supported by the CRISPR screen, which provided gene-level resolution. We fully acknowledge that additional causal genes are likely to reside within the selected regions and remain to be functionally characterized.

Impact:

Given the obvious challenges of functional testing of polygenic traits and the clear limitations of the interpretation of the results, the study will be helpful for future studies aiming to characterize polygenic traits. Unfortunately, the results are just another piece of controversial results regarding resistance against octanoic acid, a trait that is rather easy to evaluate.

The reviewer appears to imply that a trait being straightforward to phenotype necessarily implies that its genetic basis should also be straightforward to resolve. Many classic complex traits, such as human height, are simple to measure yet have an extraordinarily complex, highly polygenic genetic architecture. We believe that OA resistance represents a similar challenge: while the phenotype is readily assayed, differences in assay conditions, genetic backgrounds, and the contribution of many loci of individually modest effect make its genetic basis difficult to dissect. We have strived to be cautious in our conclusions, in particular the evolutionary interpretations; nevertheless, to our knowledge, this is the first study to provide functional evidence supporting the contribution of specific genes to OA resistance in D. sechellia, combining both loss-of-function phenotypes and expression data. As emphasized by the title of our manuscript, we view the principal contribution of this work as demonstrating how complementary experimental approaches (both of which are fairly novel for study of toxin susceptibility/resistance genetics) can be integrated to prioritize and functionally evaluate candidate genes underlying complex adaptive traits.

Recommendations for the authors:

Reviewing Editor Comments:

Both reviewers propose to include additional statistical and quantitative measures to strengthen the results. Furthermore, certain parts of the manuscript could be rephrased to make sure that the readers can understand more clearly the implications and significance of the results. To test whether the genes identified in the present manuscript (as contributing to octanoic acid resistance) are also involved in the evolution of the resistance, reciprocal hemizygosity tests using CRISPR alleles of kraken and Alkbh7 in D. simulans would be a plus, but such experiments are not required because the main focus of the paper is on the genetic basis of octanoic acid resistance, and not on evolution.

We thank the Reviewing Editor for these constructive comments. We have revised the manuscript to clarify the interpretation and significance of our findings and have addressed the reviewers’ comments throughout. Regarding reciprocal hemizygosity tests, we agree that they would provide a valuable means of assessing the evolutionary contribution of candidate genes. However, generating the necessary reagents in D. simulans represents a substantial undertaking beyond the scope of the present study, particularly given the expected modest effects of individual loci underlying this highly polygenic trait. We have nevertheless revised the end of the Results section to mention reciprocal hemizygosity tests as an important direction for future work.

Reviewer #2 (Recommendations for the authors):

(1) Provide more details about the selection tests: which sequences were used? Please report p-values. It would also be important to discuss the possibility of false positives caused by a bottleneck in D. sechellia. A genome-wide analysis could help to see if the bottleneck increased the signal of positive selection.

We are not entirely sure what additional analyses are being suggested. The sequences and methods used for the selection analyses are described in the Methods, and the statistical support for these analyses is reported in the Supplementary Material (MK test p-values and FUBAR posterior probabilities). We are also unclear as to how a genome-wide analysis would address the interpretation of the gene-specific selection analyses presented here. If the reviewer intended a different analysis, we would appreciate further clarification.

(2) The significance level of the CMH test needs to be determined with the effective population size, not with the census size, as done by the authors. This is important, as it is not clear if the candidate genes remain significant after significance adjustment based on the effective population size.

We thank the reviewer for this comment. In our analyses, effective population sizes were explicitly incorporated by Bait-ER to model the effects of genetic drift. The CMH significance thresholds, however, were obtained from neutral forward simulations following the recommended workflow for the method, which requires census population sizes rather than effective population sizes as input. To make these simulations as realistic as possible, we therefore used the observed census population sizes at each generation of the experimental evolution. Estimating generation-specific effective population sizes for use in an alternative simulation framework would require substantially more temporal data than are available here (e.g., sequencing many additional time points) and is beyond the scope of the present study.

(3) Include the allele frequency trajectory across time for the candidate genes.

We thank the reviewer for this suggestion. However, plotting allele frequency trajectories for the candidate genes is not straightforward because the evolve-and-resequence analysis identified broad linked genomic regions rather than individual causal variants. Each candidate region contains numerous SNPs spanning several kilobases, with each SNP exhibiting its own allele frequency trajectory across the 10 replicate populations. Consequently, there is no single representative trajectory for a given candidate gene, and plotting all SNPs within each region would be difficult to interpret. For kraken, however, we identified seven candidate regulatory SNPs and have plotted their individual allele frequency trajectories, which we include in Author response image 1 to illustrate the diverse patterns observed:

(4) Estimate the effect of candidate loci in the GWAS data (independent of significance).

We thank the reviewer for this suggestion. However, we are not entirely sure what analysis is being proposed. In particular, it is unclear which variants the reviewer is referring to, as our candidate genes are associated with multiple linked variants rather than a single causal SNP. Consequently, we are unsure how the effect of a candidate locus should be estimated in the GWAS dataset. Moreover, there is no guarantee that the same variants are represented in both datasets. For example, variants filtered out during the GWAS because of low allele frequency may subsequently have increased in frequency during experimental evolution and therefore contributed to the evolve-and-resequence signals. We would appreciate further clarification of the analysis the reviewer has in mind.

(5) Discuss the challenge of false positives (see: 10.1016/j.cub.2020.12.023).

We thank the reviewer for this suggestion. We agree that false positives (as well as false negatives) are an important consideration when studying complex polygenic traits, both through the initial “screening” efforts (e.g., GWAS, experimental evolution) and follow-up functional validation (e.g., RNAi, mutant, overexpression analyses). We believe this issue is already addressed in the manuscript through our discussion of the limitations of the individual approaches, the polygenic nature of OA resistance, and our cautious interpretation of the functional validation results. Our conclusions are limited to identifying candidate genes that contribute to OA resistance, rather than claiming to have identified all of the loci or variants underlying the evolution of this trait. We are therefore not sure what additional discussion the reviewer has in mind and would appreciate further clarification if a specific point from the cited study is intended.

(6) The figures with the Manhattan plots should be improved to indicate the overlapping genes in the Manhattan plots, rather than in the circle figures below. By using different colors this should be quite easy and clean.

We thank the reviewer for this suggestion. However, we believe Author response image 2 more appropriately illustrate the overlap between the CMH and BaitER analyses. Although the Manhattan plots display individual SNPs, our candidate loci are defined by broader genomic regions comprising blocks of linked significant SNPs rather than by single variants. Simply highlighting SNPs within overlapping regions would therefore add visual complexity without providing additional biological insight beyond that already captured by the circos plots. As an illustration, we provide here an example of the generation 50 Manhattan plot with overlapping SNPs highlighted in red:

(7) A more focused discussion of the assumption that functional data from D. melanogaster can explain resistance in D. simulans or D. sechellia. At some places, epistatic interactions are mentioned, but the reviewer feels that the entire screen is based on the idea that similar effects are found across species, hence, this needs to be adequately reflected in the discussion.

We thank the reviewer for raising this important point. We would like to emphasize that our study was not based on the assumption that functional effects identified in D. melanogaster or D. simulans necessarily explain the evolution of OA resistance in D. sechellia. Rather, our motivation stemmed from the longstanding difficulty of identifying individual genes underlying this highly polygenic trait using mapping approaches alone. We therefore sought to combine orthogonal experimental approaches to identify genes contributing to OA susceptibility (of D. melanogaster and D. simulans) and resistance (D. sechellia). We hoped, but did not assume, that genes supported by multiple independent lines of evidence would be informative for understanding the natural evolution of OA resistance in D. sechellia. Indeed, we acknowledge in the manuscript that the cell-based CRISPR screen, experimental evolution, and the natural evolution of D. sechellia occurred under very different selective contexts and timescales. As reflected in the title of our manuscript, our conclusions are intentionally framed around the identification of novel toxin resistance loci, while remaining cautious about their evolutionary interpretation.

(8) Discuss that the controls in the RNAi test were quite variable. Could this reflect some problems with the assay?

We do not believe that the variability among the control lines reflects a problem with the assay. Rather, the different controls (Gal4, UAS-RNAi etc.) represent distinct genetic backgrounds, each of which may exhibit a different baseline level of OA resistance. Indeed, in our recent GWAS of OA resistance (doi:10.1093/g3journal/jkag032), we observed substantial natural variation in OA resistance among D. melanogaster and D. simulans lines. Importantly, each RNAi line was compared with its corresponding genetic background control, so differences among control lines do not affect the interpretation of the individual RNAi experiments.

Version 2

Abstract

Understanding toxin resistance in insects is key to appreciate niche adaptations but remains challenging due to its often-polygenic basis. A well-known example is the specialized association of Drosophila sechellia with noni fruit (Morinda citrifolia), which is toxic to most other insects, including the closely-related Drosophila simulans and Drosophila melanogaster. Toxicity of noni is due to its high concentration of octanoic acid (OA), but the mechanisms that determine sensitivity or resistance to OA in different species remain poorly understood. Here, we experimentally-evolved D. simulans with increased OA resistance, identifying multiple loci under selection. Cross-referencing these with a genome-wide, OA-resistance CRISPR screen in a D. melanogaster cell line highlighted two proteins: Kraken, a putative detoxification enzyme expressed in digestive and renal tissues, and Alkbh7, a mitochondrial protein linked to fatty acid metabolism. Both genes show elevated expression in D. sechellia and OA-resistant D. simulans. In D. melanogaster, kraken mutants are more OA-sensitive, while Alkbh7 overexpression increased OA resistance. Importantly, mutation of these genes in D. sechellia reduced OA tolerance. Our identification of genes underlying OA resistance in laboratory and natural contexts demonstrates how complementary, cross-species selection approaches can provide insights into complex mechanisms of toxin susceptibility and adaptation; such methods could also have practical applications in the characterization of natural and artificial insecticides.

Introduction

How insects resist toxic chemicals in the environment is of both basic scientific interest – reflecting insects’ arms race with plant food sources (Erb and Reymond 2019; Jones, et al. 2022) or venomous predators (Smith, et al. 2013) – and applied interest, because, throughout history, humans have used chemicals to combat insect disease vectors and agricultural pests (Umetsu and Shirai 2020). The best-understood insecticides are those that bind to and modulate neuronal ion channels, where resistance arises through changes in these target proteins (Sparks, et al. 2020; Raisch and Raunser 2023). For example, dichlorodiphenyltrichloroethane (DDT) and pyrethroids prevent closure of voltage-gated sodium channels, resulting in neuronal excitotoxicity and organismal paralysis; diverse insect species displaying resistance for these chemicals have independently-arising mutations in genes encoding these channels (Busvine 1951; Dong 2007). Similarly, neonicotinoids bind to and hyperactivate nicotinic acetylcholine receptors, leading to neuronal death; mutations in various receptor subunits can confer resistance (Matsuda, et al. 2020). Even in such well-studied examples, however, it is thought that other neuronal targets exist (Soderlund and Bloomquist 1989) and/or that the chemicals impact receptors in non-neuronal tissues (Di Prisco, et al. 2013). For many insecticides – as well as other environmental pollutants (Gandara, et al. 2024) – the mode of action is incompletely, or not at all, characterized (Sparks, et al. 2020).

Resistance observed in nature is only partially understood, reflecting the contribution of many genes beyond those encoding target proteins, which can underlie distinct defence mechanisms (Clarkson, et al. 2018; Catteruccia 2020; Lucas, et al. 2023). For example, thickening of the cuticle to reduce entry of insecticides into the body is widely considered an effective resistance mechanism (Balabanidou, et al. 2018; Balabanidou, et al. 2019). Cuticular changes have been linked to overexpression of proteins implicated in exoskeleton development (e.g., cuticle proteins, ABC transporters) but their precise roles, and how they are upregulated, remain unclear (Balabanidou, et al. 2018). Increased expression of metabolic enzymes that degrade insecticides is another commonly observed adaptation in resistant strains, sometimes resulting from copy-number increases in the corresponding genes (Weetman, et al. 2018). Duplicated metabolic genes can also undergo neofunctionalization to produce enzymes with altered specificity, as illustrated by Cytochrome P450s in strains of the brown planthopper that recently evolved resistance to the neonicotinoid imidacloprid (Zimmer, et al. 2018), as well as by more ancient evolution of glutathione S-transferases during transition of the Scaptomyza genus of Drosophilidae to herbivory to counter toxic isothiocyanates (Gloss, et al. 2014). However, the lack of experimental resources in many insect species renders it difficult to characterize the causal contribution of specific genetic changes.

A compelling natural model clade to study the evolution of toxin resistance is the trio of closely-related drosophilid species, Drosophila melanogaster, Drosophila simulans and Drosophila sechellia (Fig. 1A). D. sechellia is an extreme specialist on the toxic noni fruit of the Morinda citrifolia shrub (Jones 2005; Auer, et al. 2021). Amongst D. sechellia’s numerous adaptations (R’Kha, et al. 1991; Dekker, et al. 2006; Auer, et al. 2020; Auer, et al. 2021; Alvarez-Ocana, et al. 2023; Shahandeh, et al. 2024), this species has evolved high resistance to octanoic acid (OA), an abundant noni chemical that is toxic to other drosophilids and more divergent insects (Legal, et al. 1994; Jones 1998; Legal, et al. 1999; Markow 2019; Auer, et al. 2021; Kaczmarek, et al. 2022; Kaczmarek, et al. 2024). Previous pioneering studies of D. sechellia’s resistance to OA has been studied through genomic mapping-based approaches in D. sechellia/D. simulans hybrids, which revealed this trait to be polygenic, comprising at least five genomic regions (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015). One of the main effect loci encompasses a cluster of 18 genes including members of the Osiris (Osi) family, which are thought to function in cuticle patterning (Ando, et al. 2019; Sun, et al. 2024). RNA interference (RNAi) knockdown of some Osi genes in D. melanogaster led to reduced tolerance of OA (Andrade Lopez, et al. 2017; Lanno, et al. 2019). However, genetic evidence for the contribution of these genes to OA resistance in D. sechellia is lacking.

Experimental evolution of D. simulans with increased OA resistance.

A Phylogeny of the drosophilid trio studied in this work. Mya: million years ago. B Schematic of octanoic acid (OA) resistance tube assay (see Materials and Methods) and selection procedure. Created with BioRender.com. C Species differences in resistance of D. simulans 04 and D. sechellia 07 strains to 3 μl OA. D Schematic illustrating the generation of the base populations for Evolve-and-Resequence (see Materials and Methods). Created with BioRender.com. E Changes in survivability to 3 μl OA of the ten D. simulans populations over generations 0-25. Box plots represent the interquartile ranges of the survivability for all the replicates; blue lines shows the generalized additive model (GAM) regression; grey shading represents the 95% confidence interval for the mean responses at each time point. F As in E for the 10 populations combined. G OA resistance for the combined 10 populations with increasing OA volumes during generation 25-31. H As in E, with 9 μl OA during generations 31-50. I As in H for the 10 populations combined. J Comparison of OA resistance of unevolved D. simulans (i.e., the original 10 outcrossed populations maintained on standard fly food; see Materials and Methods), evolved (generation 50 (G50)) D. simulans, and D. sechellia. K-N Phenotypic comparison of unevolved and evolved (G50+) D. simulans populations 1-3 for K noni resistance (G54), L fecundity (G54), M developmental viability (G57), and N lifespan (G57). For K-M, significance was assessed using the unpaired t-test correcting for multiple testing. NS p>0.05, ***p<0.001. For L-N, tests were run on standard media. No significant differences between unevolved and evolved flies were detected using the Cox regression model. See Materials and Methods for details and Data S1 for raw data.

Given the challenges of identifying candidate genes through genomic mapping-based strategies, in this work we took two complementary approaches to investigate the evolution of OA resistance in this drosophilid trio: experimental evolution in D. sechellia’s closest relative, D. simulans, through an Evolve-and-Resequence strategy (Schlotterer, et al. 2016), and CRISPR/Cas9-based genome-wide genetic screening in cultured D. melanogaster cells (Viswanatha, et al. 2018). We show how the intersection of these approaches can identify novel genes contributing to OA susceptibility/resistance in these drosophilids.

Results

Experimental evolution of OA resistance in D. simulans

A previous study demonstrated that D. simulans strains can be experimentally evolved for higher OA resistance, enabling the identification of genomic regions (through microsatellite markers) responding to selection (Colson 2004). This success inspired us to take an Evolve-and-Resequence approach, which uses high-throughput sequencing to analyse the entire genome of populations before and after exposure to specific selective pressures, thereby facilitating the identification of adaptive loci from standing genetic variation in controlled and replicable settings (Turner and Miller 2012; Schlotterer, et al. 2016; Barghi, et al. 2019; Kelly and Hughes 2019; Shahrestani, et al. 2021). For studies of insecticide resistance, it has been predominantly used to examine changes in frequency of alleles of known genes rather than as a means to identify novel loci (Zoh, et al. 2021; Sadia, et al. 2024).

The susceptibility/resistance of drosophilids to OA can be easily assessed by exposing flies to the toxin in a culture tube and counting the survivors after 24 h (Fig. 1B). In line with previous observations (Amlou, et al. 1997; Colson 2004), we found that a dose of 3 μl of pure OA (mixed with 1 ml glucose solution), corresponding to roughly half the amount estimated in 2 g of noni pulp (see Materials and Methods), leads to death of ∼60% of D. simulans but has no effect on D. sechellia (Fig. 1C). By collecting the D. simulans survivors, we reasoned that we could allow these to mate and establish the next generation (Fig. 1B), thus enriching genetic variants conferring higher resistance, similarly to (Colson 2004). Repeating the process over multiple generations and multiple replicates would allow us to track phenotypic changes in resistance and investigate the associated trajectories in allele frequencies.

Standing genetic variation should be present in the evolving populations for natural selection to act upon (Teotonio, et al. 2009; Long, et al. 2015). As the spontaneous mutation rate in drosophilids is too low for artificial selection (Keightley, et al. 2014), we generated ten independent polymorphic base populations of D. simulans from 93 isogenic strains (Signor, et al. 2018) by round-robin crossing (Fig. 1D). These lines contain millions of SNPs, but we note that they were sampled from a single wild population in California, therefore reflecting the genetic variation present in that locality. The base populations therefore might not include all variants found across the species’ range, including those in regions where D. sechellia co-occurs. To minimize bottleneck effects, each population was kept at a size of ∼1000 flies. At each generation, ∼500 flies from each population were randomly collected and selected for resistance to 3 μl OA. Over the course of 25 generations, all the populations showed increased levels of survivability, suggesting parallel adaptive evolution (Fig. 1E). Given the consistent responses across all populations, we also analysed the phenotypic changes grouping all replicates, finding that the mean survivability significantly increased from ∼43% at generation 0 to ∼78% at generation 25 (Fig. 1F).

As we kept the same amount of OA throughout the 25 generations of selection, the selective pressure lowered as the populations’ survivability increased, evidenced as a larger fraction of flies surviving and reproducing. Viewing the phenotype of populations both individually (Fig. 1E) and collectively (Fig. 1F), it was apparent that the change in phenotype plateaued at around generation 15. For continued evolution, we therefore gradually increased the amount of OA during generations 26-31 (Fig. 1G), reaching a selection pressure comparable to the initial phase of the experiment with 9 μl OA (∼38% mean survivability across all populations) (Fig. 1G). We then performed a second round of selection at this dose, observing an increase in survivability across all populations, reaching ∼70% by generation 50 (Fig. 1H). As the populations adapted to the new condition, a second plateau started around generation 40 (Fig. 1H-I). Over 50 generations we were therefore able to evolve D. simulans to have approximately 5-fold higher survivability to 9 μl OA compared to the unevolved populations (Fig. 1J), though still falling short of the survival exhibited by D. sechellia (Fig. 1J). These plateaus can be explained by the decreasing selective pressure exerted by fixed OA volumes as an ever-greater proportion of flies adapt and survive the treatment. Determining whether an upper limit exists that would constrain the evolution of a D. sechellia-like level of phenotypic resistance would require additional selection rounds using progressively higher OA concentrations over an extended (and unknown) number of generations. OA is only one component of noni fruit, although probably the major toxic chemical (Legal, et al. 1994). We therefore tested three of our evolved populations for resistance to noni fruit pulp, finding that they exhibited much higher survival on this substrate compared to the same original populations maintained under neutral conditions (i.e., no OA selection) (Fig. 1K).

Specialization of D. sechellia on toxic noni is associated with a reduced egg-laying capacity (R’Kha, et al. 1997; Markow, et al. 2009) and a shorter (albeit strain-specific) lifespan (Watada, et al. 2020; Abe, et al. 2022; Shahandeh, et al. 2024). Our success in evolving D. simulans with a partial D. sechellia-like resistance trait begs the question as to whether these selected strains exhibit other phenotypic changes as a trade-off for higher toxin resistance. We tested whether the evolved resistant populations of D. simulans exhibit similar phenotypic changes in control conditions. However, none of the three focal populations tested exhibited significant decreases in the number of eggs laid (Fig. 1L), developmental viability (Fig. 1M) or lifespan (Fig. 1N). Thus, at least within the timeframe of our experimental evolution experiment, D. simulans did not display obvious phenotypic trade-offs for the gain in OA resistance.

Dynamic directional selection of allele frequencies

We hypothesized that the observed phenotypic changes in D. simulans depended on the directional selection of alleles of consistent loci across the ten populations, while most of the genome evolves under neutrality. Allele frequencies at the selected loci should therefore consistently change across all populations, allowing us to statistically distinguish them from randomly evolving loci (Fig. 2A). We sequenced the genomes of the populations of evolved (generation 25 and generation 50) and unevolved (generation 0) flies, identifying 5,127,840 high-quality variants distributed across all chromosomes. Pairwise identity-by-state analysis indicated that as the number of generations increased, the proportions of shared alleles decreased, both between time points and between pairs of populations (left-to-right and top-to-bottom, respectively, in Fig. 2B).

Dynamic directional selection of allele frequencies during experimental evolution.

A Schematic of the analysis. B Identity-by-state analysis. The heatmap depicts the proportion of shared alleles between populations, with values ranging from 0.8 (blue, lowest shared proportion) to 1 (red, identical alleles). C,D Manhattan plot showing the probability of allele frequency changes between C generations 0 and 25, and D generations 0 and 50. Top panel shows the results of the CMH test; orange dots indicate significant variants at a threshold of-log10(p) ≥40, a conservative approximation of the top 0.000005% simulated SNPs under neutrality (Fig. S1B). Bottom panel shows the absolute log-Bayes factors estimated by Bait-ER; orange dots indicate significant variants at the conservative threshold log(0.99/0.01). Raw data in Data S2. E Circular plots illustrating overlap of predicted genomic regions under directional selection between the two statistical approaches at generation 25 (left) and generation 50 (middle), or between generation 25 and 50 (right). Raw data in Data S3.

To identify the selected genetic variants underlying the phenotypic adaptation, we analysed the genome-wide allele frequency changes both between generations 0 and 25, and between generations 0 and 50, across all populations. The identification of significant changes in frequency is non-trivial, and many methods have been developed (Vlachos, et al. 2019; Kojima, et al. 2020; Barata, et al. 2023). We therefore assessed the significance of frequency changes using two complementary approaches: the Cochran–Mantel–Haenszel (CMH) test (Cochran 1954; Mantel and Haenszel 1959), a widely-used method that tests for allele frequency changes between two time points, and Bait-ER (Barata, et al. 2023), a Bayesian method that tests for selection while explicitly modelling genetic drift. Both tests identified large portions of the genome under positive selection at generation 25 and 50 (Fig. 2C,D). After 25 generations, the two methods shared 1.90 Mb of candidate regions; the overlap encompassed 65.3% of the genomic span identified by CMH and 64.1% of the span identified by Bait-ER. After 50 generations, the two methods shared 4.77 Mb of candidate regions; here, the overlap encompassed 63.6% and 99.9% of the CMH and Bait-ER candidate regions, respectively (Fig. 2E).

Based on the analysis of genome-wide linkage disequilibrium decay (Fig. S1A), we grouped the significant SNPs identified by each of the two methods into genomic blocks according to their proximity (i.e., <50 kb). We retrieved all the candidate genes that fall within the block intervals, obtaining a total of 439 and 401 genes for the CMH and Bait-ER analysis, respectively, at generation 25, and 1238 and 174 genes at generation 50 (Data S3). These genes likely include true targets of selection, but also neighbouring genes affected by hitchhiking. The increase in CMH candidate genes over time – in contrast to the decrease observed with Bait-ER – might reflect the CMH test’s tendency to overestimate significance and to be insensitive to divergence among replicate populations.

The overlap of all significant genes between the two time points is <30%. This might reflect that a large fraction of the candidate genes results from noise caused by genetic drift and draft (hitchhiking) and/or reflect a shift in the phenotypic contribution of genes under the stronger selection (i.e., the higher dose of OA) during generations 27-50. Moreover, previous work on Drosophila using temporal sampling showed that adaptation involving complex traits can lead to non-overlapping sets of selected SNPs, in contrast with the simple expectation of continuous increases in allele frequency until fixation (Orozco-terWengel, et al. 2012; Schlotterer, et al. 2016). Gene ontology analysis is complicated at this stage as most hits are likely hitchhikers of the true causal genes. It is also hard to compare currently to previous mapping studies in D. simulans/D. sechellia hybrids (Amlou, et al. 1997; Amlou, et al. 1998; Huang and Erezyilmaz 2015) or in the earlier D. simulans experimental evolution experiment (Colson 2004), which only identified wide genomic regions. The sole exception, a 170-kb window on 3R containing only 18 genes (including several Osiris genes) (Hungate, et al. 2013) did not emerge in our analysis.

A genome-wide, pooled CRISPR screen for OA susceptibility and resistance genes in D. melanogaster S2R+ cells

Given that the previous mapping studies also contain extensive hitchhiking signals and substantial noise across large genomic regions, comparing their results with our analyses would likely yield numerous false positives and still leave a large pool of putative loci, offering limited benefit for homing in on candidate genes. We therefore sought an orthogonal screening approach. Motivated by previous observations of OA toxicity for insect Sf9 cultured cells (Kaczmarek, et al. 2022), we found that OA also leads to reduced viability of D. melanogaster S2R+ cells (Fig. 3A). This phenotype prompted us to perform a selection experiment in this cell line through exploitation of a genome-wide, pooled CRISPR/Cas9 mutagenesis approach (Fig. 3B) (Viswanatha, et al. 2018). In brief, a library of sgRNA-encoding plasmids (6-8 sgRNAs/gene) was transfected into Cas9-expressing cells where site-specific plasmid integration predominantly limits targeting to a single gene per cell. Two replicate cultures of this pool of cells were either left untreated or exposed to OA in the medium, at two different concentrations (Fig. 3B). After 26 days in culture, DNA was extracted from the cell pool, then sgRNA sequences were PCR amplified and sequenced to determine those that were enriched after OA selection (reflecting genes whose loss enhances cell viability, referred to hereafter as “susceptibility genes”) or depleted (reflecting genes whose loss leads to lower OA resistance, hereafter “resistance genes”).

A genome-wide pooled CRISPR screen for OA susceptibility and resistance genes in D. melanogaster cultured cells.

A Left: images of S2R+ cells cultured in the absence or presence of OA. Scale bar, 50 μm. Right: dose-dependent effects of OA on S2R+ cell viability, through measurement of ATP levels with a CellTiter-Glo assay (see Materials and Methods). B Schematic of CRISPR screen design. C Top 5% “susceptibility” genes (see Materials and Methods and Data S4). Box plots represent the interquartile ranges for all the replicates of the Z-scores obtained through maximum likelihood estimation to test for selection. The inset (bottom-right) displays the gene-ontology analysis of these genes. D As in C, for the top 5% “resistance” genes. E Venn diagram showing the overlap between all the genes identified in the experimental evolution at G25 and G50, and genes associated with susceptibility and resistance in the CRISPR screen. F Subportion of the top panel (CMH test) of the Manhattan plot from Fig. 2C for chromosome arms 2L and 3L, with the red arrows highlighting the regions magnified on the right showing the significant variants and linkage disequilibrium blocks (triangles) for kraken and Alkbh7 (CG14130).

To identify candidate OA susceptibility/resistance genes, we first filtered out genes essential for general cell survival (Viswanatha, et al. 2018) and genes that were not detectably expressed in S2R+ cells (using DGET (Hu, et al. 2017)). (Removal of such genes are inevitable limitations of this cell-based screening approach; see Discussion). After comparing results from each condition and replicate, we identified the top 5% susceptibility genes (Fig. 3C) and top 5% resistance genes (Fig. 3D). Amongst the 87 susceptibility genes, there was no single stand-out hit, suggesting that – at least in S2R+ cells – OA does not have a single major cellular target (unlike, for example, insecticidal bacterial toxins that bind a specific cell surface receptor (Xu, et al. 2022)). Gene-ontology analysis of these susceptibility genes revealed a significant enrichment of genes involved in membrane trafficking (Fig. 3C). As OA has previously been hypothesized to disrupt cellular membranes (Borrull, et al. 2015; Kaczmarek, et al. 2022), the recovery of these hits suggests that general reduction in membrane transport reduces OA uptake into cells and/or renders membranes less susceptible to OA treatment. 22 OA resistance genes were identified, encoding diverse types of proteins (Fig. 3D), suggesting that there are multiple mechanisms by which cells can resist the toxic effects of OA.

We next examined which of the genes identified in the S2R+ cell selection experiments might be relevant for evolution of OA resistance in whole animals by determining those also present within significant blocks of the experimental evolution analyses (Fig. 3E,F). Four susceptibility genes were identified: CG1620 (a predicted SANT-MYB domain transcription factor), CG4291 (a putative component of the splicing machinery), Nph (a chromatin remodelling factor (Emelyanov, et al. 2014)) and Debcl (a pro-apoptotic member of the Bcl-2 family (Galindo, et al. 2009)); these are all likely to have indirect roles in conferring susceptibility to OA given their broad functions. By contrast, two resistance genes have potentially direct roles in conferring tolerance to OA: kraken and CG14130. kraken encodes a putative serine hydrolase (Edwin Chan, et al. 1998) and stood out both because it was the top hit in the resistance gene screen (Fig. 3D), and because of its location in one of the most prominent peaks in the experimental evolution at generations 25 and 50 (Fig. 3F and Fig. 2C-D). CG14130 encodes the orthologue of mammalian Alkbh7, which is involved in regulating fatty acid metabolism (Solberg, et al. 2013; Zhang, et al. 2021), and is located within a significant peak at generation 25 (Fig. 3F and Fig. 2C). Therefore, intersecting the experimental evolution and CRISPR screening results allowed us to narrow down hundreds of potential candidates to two focal genes for downstream investigations.

Sequence, expression and evolutionary analyses of candidate OA resistance genes

To assess the candidacy of kraken and CG14130 (hereafter Alkbh7) in contributing to OA resistance in animals, we first examined their sequence and expression in D. melanogaster. Kraken belongs to a large family of serine hydrolases, which encompass lipases, esterases and proteases that use a nucleophilic serine residue in the active site for hydrolysis of substrates (Edwin Chan, et al. 1998). Consistent with the conservation of this serine in Kraken (Fig. 4A and Fig. S2A), an activity-based proteomic screen of the serine hydrolase superfamily in D. melanogaster provided evidence for in vivo catalytic activity of Kraken (Kumar, et al. 2021). kraken is broadly expressed in adults, with particular enrichment in the gut and Malpighian tubules (Fig. 4B), similar to previous observations of kraken expression in the embryo by RNA in situ hybridization (Edwin Chan, et al. 1998). Using the Fly Cell Atlas (Li, et al. 2022), we examined sub-tissue adult expression of kraken: in the Malpighian tubules, kraken is expressed at highest levels in the principal cells (Fig. 4C), which are involved in ion transport and detoxification. In the gut, kraken transcripts are most robustly detected in various types of enterocytes, which secrete digestive enzymes (Lemaitre and Miguel-Aliaga 2013) (Fig. 4C). However, Kraken lacks an N-terminal signal sequence (Teufel, et al. 2022), suggesting that the protein remains in the cytoplasm of these cells, unlike canonical digestive enzymes (Lemaitre and Miguel-Aliaga 2013), and more similar to detoxification enzymes of the Cytochrome P450 and Glutathione S-transferase families (Yang, et al. 2007). To examine kraken expression in vivo, we performed RNA fluorescence in situ hybridization, which detected kraken transcripts in various region of the gut (crop, crop duct, proventriculus, hindgut, ampulla) and Malpighian tubules (Fig. 4D).

Sequence and expression analyses of kraken and Alkbh7.

A AlphaFold protein models for D. melanogaster Kraken (AF-O18391-F1-model_v4) and Alkbh7 (AF-Q9VTP1-F1-model_v4) (Jumper, et al. 2021; Varadi, et al. 2022), highlighting a conserved putative catalytic serine in the active site in Kraken, and the mitochondrial targeting sequence in Alkbh7 (predicted by MitoFates (Fukasawa, et al. 2015)). See also Fig. S2A,B. B Tissue-specific expression of D. melanogaster kraken and Alkbh7 from bulk RNA-sequencing data from the Fly Atlas 2.0 (Krause, et al. 2022). C tSNE plots illustrating D. melanogaster kraken expression in single-cell transcriptomes of the Malpighian tubules and gut from the Fly Cell Atlas (Stringent 10x datasets) (Li, et al. 2022). D Left: RNAscope detection of kraken (green) transcripts in the gut and Malpighian tubules, with nuclei counterstained with DAPI (blue). Right: higher-magnification images showing kraken transcript expression in the indicated tissues. Scale bars, 200 μm. This expression pattern was observed in tissues from >20 individuals. E Expression levels of the kraken and Alkbh7 in the indicated laboratory strains (left panel) and wild caught strains (right panel) of adult female D. sechellia and D. simulans measured by qPCR (Data S6). Expression is represented as calibrated and normalized relative quantities (CNRQ). Significance was assessed using the unpaired t-test correcting for multiple testing and represented using letter codes. F Expression levels of kraken and Alkbh7 in the indicated strains and species at larval or adult stages as indicated from published RNA-seq data. (References: 1, (Watanabe, et al. 2019), 2 (Ma, et al. 2018), 3 (Kalra, et al. 2024)). Significance was assessed using the unpaired t-test correcting for multiple testing and represented using letter codes when multiple replicates were present in the original data. G Expression levels of the candidate genes at generation 0, 25 and 50 of the experimentally-evolved D. simulans populations measured by qPCR (Data S6). Significance was assessed using the paired t-test correcting for multiple testing and represented using letter codes.

Alkbh7 is a member of the alpha-ketoglutarate-dependent hydroxylase (Alkbh) family (Fig. 4A), which comprises enzymes that act on diverse substrates in processes such as biosynthesis, metabolism and epigenetic regulation (Xu, et al. 2021). As in mammals (Solberg, et al. 2013), D. melanogaster Alkbh7 displays broad tissue expression (Fig. 4B). Murine Alkbh7 is localized to the mitochondrial matrix (Solberg, et al. 2013). This property appears to be conserved in the Drosophila protein, which has a mitochondrial targeting sequence (Fig. 4A and Fig. S2B) (Fukasawa, et al. 2015); consistently, we identified Alkbh7 within D. melanogaster proteomes of the mitochondrial matrix (Chen, et al. 2015; Sen and Cox 2022).

Although the sequences of both Kraken and Alkbh7 are overall highly similar across D. melanogaster, D. simulans and D. sechellia (Fig. S2A,B), we wondered whether we could detect signs of selection at these loci in D. sechellia employing population sequence datasets of D. sechellia and D. simulans (Schrider, et al. 2018). Using the McDonald-Kreitman test (McDonald and Kreitman 1991) on the coding sequences of both genes, we found evidence for gene-level positive selection on kraken but not Alkbh7 (Fig. S2C). Because selection might act pervasively at individual sites rather than along the entire protein coding region, we also implemented a codon-based analysis using FUBAR (Fast, Unconstrained Bayesian AppRoximation) (Murrell, et al. 2013), which identified 1 and 4 codons under diversifying positive selection in kraken and Alkbh7, respectively (Fig. S2D). Although their functional relevance remains unclear, such mutations might reflect underlying evolutionary dynamics at the protein level.

We next compared the expression levels of these genes both in laboratory and wild-caught strains of D. sechellia and D. simulans (Fig. 4E), as well as in RNA-seq datasets of various strains of D. sechellia, D. simulans and D. melanogaster (Ma, et al. 2018; Watanabe, et al. 2019; Kalra, et al. 2024) (Fig. 4F, Fig. S3A,B). While there were some exceptions, overall both kraken and Alkbh7 show higher expression levels in D. sechellia adults and larvae compared to the other species. Importantly, in our experimentally-evolved D. simulans populations, we also found that both kraken and Alkbh7 were more highly expressed compared to unselected populations (Fig. 4G). The expression differences between species, and between unevolved and evolved D. simulans, reflect variation in basal levels of expression, as prior exposure to OA did not lead to changes in expression in either D. sechellia or D. simulans (Fig. S3C). RNA FISH of kraken in D. sechellia revealed similar spatial expression as in D. melanogaster, with the addition of a robust signal in the midgut (Fig. S3D). These observations suggest that elevated expression in this species is due, at least in part, to a novel expression pattern in this digestive organ.

Functional contributions of kraken and Alkbh7 to OA resistance

Given the expression differences observed in the naturally and experimentally evolved OA-resistant drosophilids, we asked whether artificially modulating gene expression would impact OA tolerance. We first performed ubiquitous RNAi of these genes in D. melanogaster (Fig. S4A) and assessed the effects on resistance to OA using a short-term, plate-based assay (Jones 1998) (see Materials and Methods). krakenRNAi led to significantly diminished OA resistance compared to genetic controls, while Alkbh7RNAidid not have a discernible effect (Fig. 5A). We next generated deletion mutants for these genes in D. melanogaster (Fig. S4B). Similar to the RNAi experiments, kraken mutants, but not Alkbh7 mutants, displayed reduced resistance to OA (Fig. 5B). All genotypes remained fully viable when no OA was added (Fig. S5A).

Functional validation of the contribution of kraken and Alkbh7 to OA resistance.

A-C Survival curves of adult flies of the indicated genotypes tested in the plate assay with 2 μl OA for kraken and Alkbh7 RNAi (A), mutants (B) and overexpression (C). Each replicate consisted of 10 2-7-day-old female flies. N replicates ≥6, n flies ≥60. Shading indicates 95% confidence intervals. Genotypes: A act-Gal4/+;UAS-krakenRNAi/+, UAS-krakenRNAi/+, act-Gal4/+, act-Gal4/+;UAS-Alkbh7RNAi/+, UAS-Alkbh7RNAi/+, act-Gal4/+ B kraken1, Alkbh71, Canton-S (wild-type genetic background control for both mutants), C act-Gal4/+;UAS-krakeno/e/+, UAS-krakeno/e/+, act-Gal4/+, act-Gal4/+;UAS-Alkbh7o/e/+, UAS-Alkbh7o/e/+, act-Gal4/+. Significance is based on the resulting mixed effect Cox regression model. NS p>0.05, *p<0.05, **p<0.01, ***p<0.001. The same data for D. sechellia are shown in both plots in C. D Survival curves of wild-type (Dsec07), and kraken (Dsec\krakenRFP) and Alkbh7 (Dsec\Alkbh7RFP) mutant D. sechellia in the plate assay with 30 μl OA. Statistics as in A. E Long-term survivability of the same genotypes as in D (n = 60 per condition) in control conditions (no OA) (left) and food supplemented with 0.9% OA (right). Statistics as in A. Raw data in Data S7.

Loss-of-function analyses might fail to uncover a role for Alkbh7 in whole animals if there are redundant resistance mechanisms absent from S2R+ cells. We therefore took a complementary approach by overexpressing these genes in D. melanogaster. Strikingly, overexpression of Alkbh7 resulted in survivability of D. melanogaster to levels close to those observed for D. sechellia under the same experimental conditions (Fig. 5C, Fig. S4C), indicating that this gene is sufficient to confer high-level OA resistance. By contrast, kraken overexpression did not have any detectable impact on OA resistance (Fig. 5C); a similar lack of effect was seen when overexpressing D. sechellia kraken (Fig. S2E). The absence of a phenotypic effect upon kraken overexpression is surprising, given its elevated expression in D. sechellia and in evolved D. simulans but this could reflect either insufficient overexpression levels (Fig. S4C) and/or genotype-by-genotype interactions that are not recapitulated when manipulating a single locus in a D. melanogaster background.

A key question is whether these genes contribute to OA resistance of D. sechellia. Implementing CRISPR/Cas9-based genome-engineering in this species (Auer, et al. 2020; Auer, et al. 2021), we generated and validated null mutants in these genes in D. sechellia (Fig. S4B). These mutant flies were homozygous viable and fertile, with no obvious morphological or behavioural abnormalities, allowing us to test their resistance to OA in the plate assay. Although we did not detect any significant difference to genetically-matched wild-type flies (Fig. 5D), this was anticipated given the polygenic nature of OA resistance in D. sechellia (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015), and the modest or absent phenotypes of equivalent mutants in D. melanogaster (Fig. 5B). Likewise, rearing D. sechellia null mutants on noni pulp (Fig. S5B), we observed no significant differences between mutant and wild-type flies over >30 days, indicating that loss of either gene alone does not compromise D. sechellia’s capacity to persist in its natural host fruit.

We reasoned that more subtle differences might be detected by monitoring D. sechellia survivability at higher OA concentrations over several days. We therefore exposed flies to food containing 0.9% OA, approximately threefold higher than the ∼0.3% OA reported for ripe noni fruit (3.06 g OA/kg fruit) (Pino, et al. 2009). Indeed, we observed higher mortality rates of both kraken and Alkbh7 mutants compared to genetically matched wild-type flies over a period of 6 days (Fig. 5E). By contrast, both mutants survived equally well as controls in the absence of OA over this time period (Fig. 5E). To determine whether the observed OA sensitivity was attributable to a general reduction in fitness, we examined survivorship under several conditions of thermal stress (Fig. S5C): mutant and wild-type flies showed no significant differences across all temperatures tested, indicating that the increased mortality observed upon OA exposure does not result from an overall physiological weakness of the knockout lines. These data provide the first evidence for the contribution of specific genes to OA resistance in D. sechellia. However, future reciprocal hemizygosity tests (Stern 2014) and/or allelic swap studies with D. simulans or D. melanogaster would be required to assess if and how they contribute to species-specific differences in OA sensitivity (e.g., through structural and/or regulatory differences in the genes).

Discussion

Identifying specific loci underlying phenotypic evolution is challenging due to the polygenicity of the vast majority of adaptive traits, with recent advocacy of new experimental paradigms to bridge genotype-phenotype relationships (Tautz, et al. 2026). Investigating the evolution of chemical toxicity is an attractive model trait because of the relatively easy phenotypic read-out. To gain insight into the genetic basis of susceptibility and resistance to a natural toxin, OA, we have intersected the results of selection over vastly different timescales: weeks in cell lines, years in experimentally-evolved laboratory strains, and millennia during speciation.

Unlike insecticides that activate or antagonize broadly-expressed neuronal ion channels, the mode of action of OA appears to be mechanistically very different. Beyond insect cell lines (our work and (Kaczmarek, et al. 2022)), OA is also toxic to budding yeast (Mota, et al. 2024) and bacteria (Liu, et al. 2013), suggesting that this toxin acts via a more fundamental cellular mechanism. Indeed, a genome-wide, knock-out screen in yeast for genes required for tolerance of OA highlighted contributions of several core cellular processes, such as vesicle trafficking, mitochondrial function or chromatin regulation (Mota, et al. 2024), reminiscent of our CRISPR screen results. Consistent with this idea, the results of our experimental evolution and CRISPR screening, as well as earlier mapping efforts in D. sechellia/D. simulans hybrids (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015) revealed multiple genomic regions/genes contributing to the resistance to OA.

Considered separately, the experimental evolution and cell-based CRISPR screens identified many candidates. That there was very limited overlap between the screen results is unsurprising given their very different designs. The loss-of-function, cell-based screen of course has several limitations, as it cannot identify genes that are not expressed in the cell line, those essential for cell viability in the absence of OA, nor genes that function in biological processes only pertinent at a tissue/organismal level (e.g., cuticle biogenesis, neuronal signalling). Nevertheless, both lists are likely to contain multiple genes relevant to understanding OA susceptibility and resistance, and hence mode(s) of action of OA. Here, we focused on two resistance gene candidates, kraken and Alkbh7, which were common hits of both screens. We note, however, that neither of these genes were identified in previous studies of OA resistance (Amlou, et al. 1997; Amlou, et al. 1998; Jones 1998; Huang and Erezyilmaz 2015), which could reflect several factors, such as differences in OA toxicity assay conditions, developmental stages, species and molecular readouts (DNA or RNA), and the evident multilayered, polygenic nature of this trait (discussed further in (Marconcini, et al. 2026)). We note, however, that our recent GWAS of OA resistance in the founder panel of D. simulans used to establish the experimental evolution populations identified bez and CG13003 as candidate genes (Marconcini, et al. 2026), which showed significant CMH signals at generation 25 (but not generation 50) of the evolve-and-resequence experiment. (Neither gene could have been recovered in the CRISPR screen because they are not expressed in S2R+ cells (Hu, et al. 2017)).

The (partial) necessity, but insufficiency, of kraken, and lack of necessity but sufficiency of Alkbh7 to confer OA resistance implies fundamentally different roles of these proteins. The expression of kraken in the gut and renal system supports a hypothesis of this protein as a detoxification enzyme, although future work will be necessary to determine whether Kraken has a direct role in OA degradation. The mild loss-of-function phenotype of kraken in both D. melanogaster and D. sechellia likely reflects substantial redundancy of defence mechanisms, including other potential detoxification mechanisms, especially given the plethora of enzymes expressed in the fly gut and Malpighian tubules (Lemaitre and Miguel-Aliaga 2013; Dow, et al. 2022). Indeed, extremely few associations between such enzymes and substrates and/or physiological roles have previously been defined. One exception is the Cytochrome P450 gene Cyp6g1, which has a key role in Malpighian tubules in conferring resistance to DDT (Yang, et al. 2007). Our data on Kraken therefore provide one of the first links between a natural toxic chemical and a putative detoxification enzyme.

The mitochondrial Alkbh7 enzyme likely has a distinct role in OA resistance. Insights from mammals are informative: mice lacking Alkbh7 are obese, with higher body fat levels, proposed to result from a defect in metabolism of fatty acids (Solberg, et al. 2013). Molecularly, one study suggests that Alkbh7’s metabolic role stems from its regulation of RNA editing to control mitochondrial protein translation (Zhang, et al. 2021), although it is unclear whether this is its only function. While the Alkbh7 loss-of-function phenotype is mild in drosophilids, it is striking that artificially elevated expression in D. melanogaster is sufficient to produce D. sechellia-like levels of OA resistance. This result implies a potent effect of Alkbh7 on the mitochondrial metabolic network to modulate toxin resistance, perhaps through control of fatty acid metabolism. While such ideas can be tested in future studies, we note that defects in lipid metabolism and mitochondrial function have been observed upon exposure to other types of insecticide (Martelli, et al. 2020; Martelli, et al. 2022). Intriguingly, our implication of Alkbh7 in the niche adaptation of D. sechellia has a potential parallel in bats, where this gene displays evidence of positive selection during dietary diversification of different species (Gutierrez-Guerrero, et al. 2020).

Taken together, our work has demonstrated the power of orthogonal selection approaches in the laboratory to understand toxin resistance, which might be relevant to understand a historic paradigm of this phenomenon (R’Kha, et al. 1991), as well as offering clues as to how the natural toxin OA exerts its lethal effects. Moreover, the methodology – notably cell-based CRISPR screening, which can now be extended to both gain-of-function approaches (Xia, et al. 2023) and mosquito cell lines (Viswanatha, et al. 2021) – is applicable to explore the molecular mechanisms of susceptibility and resistance of insects to the wide-range of artificial toxic chemicals used in the environment (Gandara, et al. 2024). Such knowledge could have application in the future characterization of novel types of insecticides.

Materials and Methods

Drosophila strains and husbandry

Drosophila stocks were maintained on standard wheat flour–yeast–fruit–juice medium under a 12 h light:12 h dark cycle at 25°C. Wild-type strains, as well as mutant and transgenic lines used in this study are listed in Table S1. Experimental evolution was performed using a subset of published D. simulans isogenic strains (Signor, et al. 2018) (Table S2).

Generation of mutant and transgenic flies

Loss-of-function mutants

we generated D. melanogaster mutants for kraken and Alkbh7 using in vivo CRISPR/Cas9 mutagenesis (Port, et al. 2020) by crossing flies expressing Cas9 in the germline (nanos-Cas9) to those expressing sgRNAs targeting the desired gene. For kraken, we used P{hsFLP}1,y1,w1118;P{HD_CFD02326}attP40 as a source of sgRNAs driven by Act5C-GAL4; for Alkbh7, we used the CG14130 sgRNA line described in (Lenče 2021). Mutants were validated by genomic PCR and sequencing (Fig. S4B).

D. sechellia mutant flies were generated by WellGenetics Inc. using modified methods of (Kondo and Ueda 2013). For kraken (GM16772/LOC6617307) the upstream sequence CTGTAATGCTGGACGCGGAT[TGG] (PAM in square brackets) and the downstream sequence CCTGGTCACCCCGGATCGAG[TGG] were cloned separately into a U6 promoter plasmid. The 3xP3-RFP cassette, which contains a floxed 3xP3-RFP flanked by a 5’ homology arm (999 bp: from-1043 nt to-45 nt relative to the kraken ATG) and a 3’ homology arm (1063 bp: +1388 nt to +2450 nt) were cloned into pUC57-Kan as a donor template for repair. For Alkbh7 (GM25300/LOC6605344) the upstream sequences ACTAATAAGAGTCGACCGGG[TGG], AACAGTCAAACAGCTGTTCC[TGG] and downstream sequences GTCCCAGTTTCAGGGTACGC[TGG], ATGGAAATGCGACGTGTCCT[GGG] were cloned separately into a U6 promoter plasmid. The 3xP3-RFP cassette, which contains a floxed 3xP3-RFP, 5’ homology arm (1037 bp:-1121 nt to-85 nt relative to the Alkbh7 ATG) and 3’ homology arm (964 bp: +940 nt to +1903 nt) were cloned into pUC57-Kan. DNA plasmids encoding the sgRNAs and the corresponding homology-directed repair template were microinjected into embryos of Dsec.07 pBAC(nos:Cas9,3P3-YFP) (Auer, et al. 2020). In both cases, a single F1 line expressing the 3xP3-RFP selection marker was obtained, which was validated by genomic PCR and sequencing (Fig. S4B).

D. melanogaster and D. sechellia mutant flies were backcrossed for six generations to Canton-S and Dsec07, respectively.

Transgenic flies: to make the UAS-Dmel\kraken and UAS-Dsec\kraken transgenes, we amplified (using oligonucleotides listed in Table S6) the regions from the start to the stop codons from w1118 and Dsec07 genomic DNA and subcloned them into a pUASTattB vector (Bischof, et al. 2007). Transgenic flies were obtained by phiC31 integrase-mediated transgenesis into attP2 performed by BestGene Inc. and verified by sequencing.

Toxin resistance assays

Tube assay

resistance to OA for the experimental evolution setup was scored adapting a previous protocol (Amlou, et al. 1997; Colson 2004). In brief, flies were lightly anesthetized with CO2 and ∼50 animals were placed in standard cultured tubes (25 mm diameter × 95 mm height, Milian SA) and provided with 1 ml of a 3% glucose solution on a tissue paper (Kimtech, 7552). 3 μl of pure OA (Sigma-Aldrich, CAS 124-07-2, C2875) were directly pipetted onto the tissue while the flies remained anesthetized. Direct contact with the acid was scrupulously avoided until all flies recovered. The fraction of alive flies was scored after 24 h. For the experimental evolution experiment, the surviving flies were collected (see below).

Plate assay

to measure OA resistance in the genetically-manipulated D. melanogaster and D. sechellia strains, we used a plate assay (Jones 1998). Compared to the experiments in tubes, this assay offers greater temporal resolution and higher dynamic range (in contrast to the experimental evolution experiment, we did not strive to maintain a certain fraction of surviving flies), as well as greater statistical power from tests of fewer animals. Ten female flies were lightly anesthetized with CO2 and placed in a Petri dish (60 mm diameter × 15 mm height) in which 2 μl of pure OA (or 30 μl for D. sechellia) were spotted in the centre of the underside of the lid. The flies were allowed to recover (typically <5 min), and immobile/dead were scored every 5 min for 1 h. The survival curves were calculated and plotted using the R package survminer (10.32614/CRAN.package.survminer). The statistical analysis was conducted fitting a mixed effect Cox regression model as implemented in the R package coxme (Therneau 2024). Genotype was included as a fixed effect, and replicate dishes were modelled as a random effect.

Noni resistance assay

noni fruit was collected from M. citrifolia plants grown in the University of Lausanne greenhouses. Five ripe (pale yellow) fruits were homogenized together, and the pulp puree frozen at-20°C. The noni assay was performed with the same setup as for tube assay described above, using 2 g of thawed noni puree placed on the bottom of the tube. Noni contains 3.06 g OA/kg fruit (Pino, et al. 2009); assuming an equivalent quantity in our fruits, 2 g noni pulp should contain 6.73 µl OA, approximately double the amount used for the experimental evolution between generations 1 to 25, and two thirds of the concentration during generations 30-50.

Long-term assay

To assess the long-term effects (i.e., beyond 24 h) of OA or noni pulp, two-day-old mated female flies were individually housed in vials containing either food supplemented with OA or 2g of noni puree. OA-supplemented food was prepared by thoroughly mixing OA into Formula 4-24® Instant Drosophila Medium, Blue (Carolina Biological Supply) to a final concentration of 0.9%. Flies were flipped into fresh food every two days, and mortality was monitored daily. Statistical analysis was run as for the plate assay.

Fecundity and developmental viability

10 mated adult female flies (4-7 days old) were allowed to lay eggs for 24 h on agar plates (2% agar, 2.5% sucrose in a 1:3 grape juice:water mix). The total number of eggs on each plate was counted. From each plate, 100 eggs were collected and transferred to standard fly food. Emerging adults were counted and removed daily until no further flies emerged.

Lifespan

Two-day-old mated female flies were individually housed in vials containing standard fly food and monitored daily until death. Fresh food was provided every 2–3 days.

Experimental evolution

Selection procedure: to generate the base populations for experimental evolution, we used 93 isogenic lines (Table S2) of the D. simulans panel established by (Signor, et al. 2018). We first crossed each of the lines together following a round-robin design to ensure equal contribution of each line to the starting genetic pool; 5 males and 5 females from each cross were then combined in ten independent replicates (Fig. S1D). These ten populations were maintained in bottles (57 mm diameter ×103 mm height, Milian SA) of standard fly food at high population size (∼1000 individuals per population) to minimize the effect of genetic drift. Each population was subdivided into smaller groups of ∼250 individuals per bottle (i.e., four bottles per population) to reduce competition. At each generation, individuals from the four bottles were mixed to prevent bottle-specific genetic drift. The experimental evolution experiment was initiated 10 generations after the creation of these populations.

From each population, 500 individuals (2-7 day old) were randomly collected and further split into 10 groups of 50 mixed-sex flies (i.e., ∼5000 flies total). Each group of 50 flies was then placed under selection with 3 μl of OA for 24 h (as described above); the survivors were collected and allowed to breed in food vials for 6-8 days, transferring them to fresh vials every 2-3 days. In the next generation, 500 flies per population were collected only from the last seeded vial, to maximize the probability of these flies being parented by the OA-resistant males and females and not offspring generated by females that mated before the selection procedure. These flies were then subjected to the same selection procedure. This process was repeated for 25 generations, and survivability was scored at each generation.

A second round of experimental evolution was performed by increasing the amount of OA to 9 μl between generations 31 and 50; this amount was determined empirically during generation 26-31 to re-establish the survivability close to the initial level (Fig. 1G). All successive generations were maintained following the same procedure.

As a control, the ten original populations were maintained in parallel in the same conditions as previously described (i.e. ∼1000 individuals per population subdivided into smaller groups of ∼250 individuals per bottle).

DNA sequencing and analysis: we pooled ∼200 individual flies per population per time point (generations 0, 25 and 50). Genomic DNA (and total RNA for qPCR, as described below) extraction was performed using the Quick DNA/RNA Miniprep plus kit (Zymo Research, D7003) following the manufacturer’s instructions. Samples were stored at-80°C.

We performed paired-end whole genome sequencing on an Illumina NovaSeq 6000 following Nextera DNA flex preparation, at an average coverage of >200×. Read quality was assessed with Fastqc (Wood 2011) and adapters were trimmed with Trimmomatic (Bolger, et al. 2014) providing the Nextera adapters sequences. We aligned the trimmed reads to the D. simulans reference genome (NCBI GCF_016746395.2) using bwa-mem (Li 2013) and the results were sorted and compressed using samtools (Danecek, et al. 2021). Duplicate reads were then removed with Picard MarkDuplicates (https://broadinstitute.github.io/picard/). Unplaced scaffolds and alignments with mapping quality <20 were removed with samtools. We then combined all the aligned sequences into a samtools pileup file and split this into individual chromosomes. Regions with zero coverage were removed with custom scripts. Sync files for each chromosome were generated with the mpileup2sync.pl script implemented in Popoolation2 (Kofler, et al. 2011) and used as input files for all the Evolve-and-Resequence analyses.

To explore the genetic variation of our samples we generated vcf files with vcftools 0.1.17 (Danecek, et al. 2011) and binary files with plink1.9 (Purcell, et al. 2007; Purcell 2020). We calculated the identity-by-state (IBS) matrix as implemented in plink1.9 (function-distance ‘ibs’).

Evolve-and-Resequence analysis: allele frequency changes between generation 0 and generations 25 or 50 were investigated via Popoolation2 and Bait-ER (Barata, et al. 2023).

We performed a Cochran–Mantel–Haenszel (CMH) test via the cmh-test.pl script implemented in Popoolation2 comparing base and evolved populations with parameters --min-count 12 and --min-coverage 50. We estimated a plausible significance threshold approximating the top 0.000005% of the simulated p-values based on the effect of genetic drift alone by running genome-wide forward simulations under neutrality in mimicree2 (Vlachos and Kofler 2018). We used the same genomic information from the original 93 lines as input and the average real survivability ratios observed throughout the selection experiment as –selection-regime parameter and 500 as –population-size parameter (Fig. S1B).

For Bait-ER, the input sync files were filtered to match --min-count 12 and --min-coverage 50 and scaled to a uniform coverage of 100 as in (Vlachos, et al. 2019) to avoid numerical problems due to high coverage. We ran the algorithm on each chromosome separately. Bait-ER models the effect of genetic drift directly using estimates of the effective population size (Ne). We calculated Ne estimates for each chromosome in each replicate population with the R package poolSeq (Taus, et al. 2017) and used the median value of each chromosome as “Population_size” parameter (Table S3). Other parameters included “Sampling correction” = 0 (i.e., binomial) and “Prior parameters” =0.001 (shape), 0.001 (rate). Data points were considered significant when the absolute natural logarithm of the Bayes Factor exceeded the conservative threshold of log(0.99/0.01) as in (Barata, et al. 2023).

The linkage disequilibrium decay in the 10 evolved populations was analysed across the whole genome using PopLDdecay (Zhang, et al. 2019). Linkage disequilibrium maps of focal regions were generated with LDBlockShow (Dong, et al. 2021) using the Solid Spine method based on D’ statistics originally introduced in Haploview (Barrett, et al. 2005).

For the genome-wide screen, all the genes within selection blocks were considered potential candidates. Based on the results on linkage disequilibrium decay, a block was arbitrarily defined as a genomic interval where at least two significant SNPs were found at a distance less than 50 kb. Consecutive SNPs respecting this definition were grouped into larger blocks and visualized as circus plots using the R package circlize (Gu, et al. 2014). The genes overlapping such blocks were extracted using BEDtools intersect (Quinlan and Hall 2010). Orthologous genes in D. melanogaster were retrieved using a combination of the OMA browser (Altenhoff, et al. 2021), DAVID (Huang da, et al. 2009; Sherman, et al. 2022), BLAST (Camacho, et al. 2009), NCBI (Sayers, et al. 2025) and FlyBase (Ozturk-Colak, et al. 2024), and subsequent manual curation.

Genome-wide CRISPR/Cas9 screen in D. melanogaster S2R+ cells

Library design, construction, and transfection: design, synthesis and cloning of the genome-wide sgRNA library are previously described (Viswanatha, et al. 2025). To determine sub-lethal doses of OA appropriate for screening, the effect of a gradient of OA supplementation into media on attP+ Cas9+ cells was determined using CellTiter-Glo (Promega) following the manufacturer’s protocol; luminescence measurements were made using a Spectramax Paradigm (Molecular Devices).

For the screen, attP+ Cas9+ D. melanogaster S2R+ cells (RRID:CVCL_UD30) were transfected using Effectene and a 1:1 molar ratio of sgRNA library and pIntAC, as described (Viswanatha, et al. 2025). Briefly, transfected cells were distributed in 5 ml aliquots across ten 100 mm culture dishes and incubated overnight, then 5 ml of fresh media was added to each plate. After 4 days, cells were subjected to puromycin selection for 12 days, subculturing every 4 days, to establish a stable sgRNA-expressing cell library. For the selection assay, 2.5 × 107 cells were plated onto four 15 cm diameter cell culture dishes to ensure sufficient sgRNA coverage, with each sgRNA represented approximately 1000 times (Viswanatha, et al. 2025).

OA selection assay: library-containing cells were left untreated or treated with either 0.075 μg/μl or 0.05 μg/μl OA (Sigma-Aldrich, CAS 124-07-2, C2875). Each replicate consisted of four 15 cm diameter cell culture dishes seeded with cells, achieving ∼1000× coverage. Two replicate sets were included for each of the two OA treatment conditions and one replicate for the untreated control. During the selection period, cells were split 1-2 times per week to avoid overgrowth. An aliquot of cells from each replicate was collected and stored for genomic prep following each split and at the end of the assay. The final samples were collected after 26 days of OA or control treatment.

Genomic DNA extraction and sequencing: genomic DNA was extracted from samples collected during the screen and at the final endpoint using a Zymo Quick-gDNA Miniprep kit (Zymo Research, D3025). DNA fragments containing the sgRNA sequences were amplified by PCR using what we calculated to be at least 1000 genomes per sgRNA as templates for each sample. PCR fragments were in-line barcoded using a previously described approach (Viswanatha, et al. 2018; Viswanatha, et al. 2019; Viswanatha, et al. 2025). Next-generation sequencing (Illumina NextSeq) was performed at the Biopolymers Facility at Harvard Medical School (RRID:SCR_007175). FASTQ sequence files were demultiplexed using in-line barcodes as sequence tags as described previously (Viswanatha, et al. 2018; Viswanatha, et al. 2025). All further processing was done using MAGeCK (v.0.5.4 and 0.5.9.4) (Li, et al. 2014) to generate readcount files and perform robust rank aggregation (RRA) analysis and Maximum likelihood estimation directly on the generated readcount files using default parameters. Plasmid readcounts were used to normalize RRA values.

After removing essential genes at a 2% FDR (2,517 genes; Data S4), we considered as candidates the top 5% genes overlapping across the two conditions that were over-represented or under-represented in the OA-exposed cell populations compared to the control cells (i.e., representing “susceptibility” and “resistance” genes, respectively). The gene-set enrichment analysis was performed with PANGEA 1.1 Beta using the Drosophila GO subsets (GO slim) and a p-value cut-off of 0.05 (Hu, et al. 2023).

qPCR analysis

Reverse transcription of total RNA (extracted as described above) was performed with the PrimeScript first-strand cDNA synthesis kit (Takara, 6110A) following the manufacturer’s instructions. Samples were stored at-80°C.

Quantitative PCR experiments were conducted on a QuantStudio 6 Flex Real-Time PCR System using the PowerUp™ SYBR™ Green Master Mix following the manufacturer’s instructions. Each sample consisted of at least three biological replicates. We performed three technical replicates per reaction and removed replicates where the standard deviation was equal or greater than 0.5. We determined target specific amplification efficiencies for each gene (Table S4, Table S6). Relative quantification and statistical analysis were performed with the software qbase+ v3.4 (Hellemans, et al. 2007) using Calibrated Normalized Relative Quantities (CNRQs). Normalization was performed on the geometric mean of the relative quantities for the three most commonly-used reference targets: Act5C, αTub84B and RpL32 (Lu, et al. 2018). We assessed reference target stability by the geNorm expression stability value (M) and the coefficient of variation of the normalized reference gene relative quantities (CV) (Table S5) (Vandesompele, et al. 2002; Hellemans, et al. 2007).

Population genetics analysis

Publicly available WGS data for 41 D. sechellia and 20 D. simulans genomes were downloaded from the Sequence Read Archive (PRJNA215932 and PRJNA395473) (Rogers, et al. 2014; Schrider, et al. 2018) with fasterq-dump (Leinonen, et al. 2010). FASTQ files were processed and aligned as described in the “Experimental evolution” section. Aligned files were phased with samtools phase (Danecek, et al. 2021) and the genes of interest were extracted with pilon (Walker, et al. 2014) to find variation among populations. CDS sequences were aligned based on amino acid information using translatorX (Abascal, et al. 2010). Gene sequences were aligned using Clustal Omega (Madeira, et al. 2024). McDonald-Kreitman test, Tajima’s D and Fu and Li’s statistics were calculated in DNAsp 6 (Rozas, et al. 2017). FUBAR analyses (Fast, Unconstrained Bayesian AppRoximation) were performed in HyPhy 2.5 (Kosakovsky Pond, et al. 2020). The consensus sequences of D. simulans and D. sechellia were computed by SnapGene (www.snapgene.com) “export consensus” with a >50% threshold. The sequence alignment and secondary structures were depicted using ESPript 3 (Robert and Gouet 2014).

RNA fluorescence in situ hybridization

To detect kraken mRNA in situ, we used the RNAscope Multiplex Fluorescence Detection Reagents v2 (Advanced Cell Diagnostics, 323110) and RNAscope H2O2 and Protease Reagents (Advanced Cell Diagnostics, 322381). Surfaces and equipment were cleaned with RNAzap (Invitrogen, AM9780) prior to dissection. Intestinal tissues were dissected from D. melanogaster (Canton S) and D. sechellia (D. sechellia 07) females in RNAlater (Sigma-Aldrich, R0901) diluted 1:50 in DEPC-treated water and briefly rinsed in 1× PBS (prepared in ultrapure water) (ThermoFisher Scientific, 70011044). Dissected tissues were mounted on poly-L-lysine-coated slides in 1× PBS and fixed immediately in cold 4% formaldehyde (Sigma-Aldrich, F8775) for 30 min. The fixative was removed, and tissues were incubated with few drops of hydrogen peroxide for 10 min at room temperature (RT). Tissues were then washed 3× 5 min in wash buffer (Advanced Cell Diagnostics, 320058) preheated to 40°C. For tissue dehydration, slides were incubated with 200 μl of ethanol solutions in the following order: 50% ethanol for 5 min at RT, 70% ethanol for 5 min at RT and 100% ethanol for 5 min at RT (repeated twice). Slides were air-dried for 5 min, and samples incubated with few drops of Protease III for 30 min at RT.

To hybridise probes, tissues were washed in 1× PBS for 3× 5 min. 50 μl of C1 (D. melanogaster-kraken-RA (NM_057917), Advanced Cell Diagnostics, 1850601-C1); and 1 μl of C2 (D. melanogaster-Rpl32 (NM_079843.4), Advance Cell Diagnostics, 533451-C2) RNAscope FISH probes were preheated to 40°C for 10 min, cooled at RT for 5 min, applied to tissues and the slides were placed in a humidified chamber for hybridization at 40°C for 2 h. Slides were hybridized with AMP reagents in sequential steps. A few drops of AMP1, warmed to RT, were applied to slides, which were incubated at 40°C for 30 min, followed by washes in wash buffer 3× 5 min. Next, a few drops of AMP2, warmed to RT, were applied to slides and incubated at 40°C for 30 min, followed by washes in wash buffer 3× 5 min. Finally, a few drops of AMP3, warmed to RT, were applied to slides, and incubated at 40°C for 15 min, followed by washes in wash buffer 3× 5 min.

A few drops of HRP-conjugated probe was applied for C1 and slides were incubated at 40°C for 15 min. After, rinsing in wash buffer 3× 5 min, Opal dyes 520 nm (Akoya Biosciences, OP-001001) and/or 570nm (Akoya Biosciences, OP-001003 (1:1000 in TSA multiplex buffer, Advanced Cell Diagnostics, 322809) were applied to slides and incubated at 40°C for 30 min. Slides were then washed and treated with HRP blocker (a few drops per slide) for 15 min at 40°C and washed in wash buffer for 3× 5 min. Tissues were counterstained with DAPI for 5 min at RT and washed once in the wash buffer, before mounting in Vectashield (Vector Laboratories, H1000).

A Leica STELLARIS-DIVE confocal microscope was used to generate all confocal images, using a water-immersive 40× objective or an oil-immersive 63× objective. The images were acquired using both HyDs as well as PMTs tailored for the fluorophores of each sample.

Statistics and reproducibility

Data were analysed and plotted in R 4.0.3 (R Core Team 2021).